# Comparing the statistical power of $F_{ST}$ versus $f_3$

Let's look at the equations behind $F_{ST}$ and $f_3$ statistics as measures
of differentiation between two populations:

Based on these equations, would you expect one of them to be more robust to
$N_e$ differences between those populations? In other words, if you were
interested in measuring how far in the past the populations began
differentiating, would both of them give you a reliable answer, regardless of
the $N_e$ of those populations?

Let's confirm this using simulations again.

```{r}
#| message: false
library(slendr)
init_env()

library(dplyr)
library(ggplot2)
library(purrr)
```

```{r}
outgroup <- population("outgroup", N = 1000, time = 1)
X <- population("X", N = 1000, time = 3000, parent = outgroup)
Y <- population("Y", N = 1000, time = 3000, parent = outgroup)

model <- compile_model(list(outgroup, X, Y), generation_time = 1, simulation_length = 10000)

plot_model(model, proportions = TRUE)

ts <- msprime(model, sequence_length = 10e6, recombination_rate = 1e-8) %>%
  ts_mutate(mutation_rate = 1e-8)
```

```{r}
samples <- ts_names(ts, split = "pop")

value_fst <- ts_fst(ts, sample_sets = samples[c("X", "Y")])$Fst
value_f3 <- ts_f3(ts, A = samples["outgroup"], B = samples["X"], C = samples["Y"])$f3

result <- data.frame(stat = c(value_f3, value_fst))
result$Ne <- 1000

result
```

```{r}
simulate_fst_f3 <- function(Ne) {
  outgroup <- population("outgroup", N = Ne, time = 1)
  X <- population("X", N = Ne, time = 3000, parent = outgroup)
  Y <- population("Y", N = Ne, time = 3000, parent = outgroup)
  
  model <- compile_model(list(outgroup, X, Y), generation_time = 1, simulation_length = 10000)
  
  plot_model(model, proportions = TRUE)
  
  ts <- msprime(model, sequence_length = 10e6, recombination_rate = 1e-8) %>%
    ts_mutate(mutation_rate = 1e-8)
  
  samples <- ts_names(ts, split = "pop")
  
  value_fst <- ts_fst(ts, sample_sets = samples[c("X", "Y")])$Fst
  value_f3 <- ts_f3(ts, A = samples["outgroup"], B = samples["X"], C = samples["Y"])$f3

  result <- data.frame(statistic = c("Fst", "f3"), value = c(value_fst, value_f3))
  result$Ne <- Ne
  
  result
}
```

```{r}
Ne_steps <- c(100, seq(from = 500, to = 10000, by = 500))
Ne_steps
```

```{r}
result <- map_dfr(Ne_steps, simulate_fst_f3)
```

```{r}
ggplot(result, aes(Ne, value, color = statistic)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ statistic, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45)) +
  ggtitle("Fst and f3 for X-Y split at 3000 generations ago, assuming given Ne")
```

Conclusion: unlike $F_{ST}$, the $f_3$ statistic appears to be an unbiased
estimator of the population divergence with respect to $N_e$ of the two
lineages!

