# Tree-sequence statistics

## Introduction

In this session, you are going to build on top of the results from the previous
[session on programming demographic
models](https://bodkan.net/simgen/popgen-models.html). Specifically, you will
learn how to compute population genetic statistics on _slendr_-simulated tree
sequences using [its
interface](https://bodkan.github.io/slendr/reference/index.html#tree-sequence-statistics)
to the _tskit_ Python module.

## Exercise 0 --- catching up

**First, create a new R script `popgen-stats.R` and paste in the following
code.** This is one of the possible solutions to Exercise 1 of the session on
building demographic models. **It's easier if we all use it to be on the same
page from now on, starting from the same model and the same simulated tree
sequence:**

```{r}
#| collapse: true
#| output: false
library(dplyr)

library(slendr)
init_env(uv = TRUE)

chimp <- population("CHIMP", time = 7e6, N = 5000)
afr <- population("AFR", parent = chimp, time = 6e6, N = 15000)
eur <- population("EUR", parent = afr, time = 60e3, N = 3000)
nea <- population("NEA", parent = afr, time = 600e3, N = 1000, remove = 40e3)

gf <- gene_flow(from = nea, to = eur, proportion = 0.03, start = 55000, end = 50000)

model <- compile_model(
  populations = list(chimp, nea, afr, eur),
  gene_flow = gf,
  generation_time = 30
)

# sampling of two Neanderthals at 70kya and 40kya
nea_samples <- schedule_sampling(model, times = c(70000, 40000), list(nea, 1))

# sampling one Chimpanzee sample, 5 African samples, and 10 European samples
present_samples <- schedule_sampling(model, times = 0, list(chimp, 1), list(afr, 5), list(eur, 10))

# sampling one European sample between 50kya and 2kya, every 2000 years
times <- seq(40000, 2000, by = -2000)
emh_samples <- schedule_sampling(model, times, list(eur, 1))

# binding all sampling schedules together
schedule <- rbind(nea_samples, present_samples, emh_samples)

# simulate a tree sequence with mutations (feel free to increase the amount
# of sequence from 20 Mb to something like 100 Mb if you want)
ts <-
  msprime(model, sequence_length = 20e6, recombination_rate = 1e-8, samples = schedule, random_seed = 1269258439) %>%
  ts_mutate(mutation_rate = 1e-8, random_seed = 1269258439)
```

As a sanity check, let's use a couple of _tidyverse_ table-munging tricks to
make sure the tree sequence does contain a set of samples which matches our
intended sampling schedule (particularly the time series of European
individuals and the two Neanderthals):

```{r}
# total number of recorded individuals in the tree sequence
ts_samples(ts) %>% nrow

# times of sampling of each recorded Neanderthal
ts_samples(ts) %>% filter(pop == "NEA") %>% pull(time)

# count of individuals in each population
ts_samples(ts) %>%
  mutate(age_group = if_else(time == 0, "present-day", "ancient")) %>%
  group_by(pop, age_group) %>%
  summarize(n = n()) %>%
  select(age_group, pop, n)
```

::: aside

**Note:** These bits of _tidyverse_ code are extremely helpful when you're working
with large tree sequences with many individuals as sanity checks that your
sampling worked as intended. I'm listing them here in case you've never worked
with the _tidyverse_ family of R packages before (such as the _dplyr_ package
where `filter()`, `group_by()`, `tally()`, and `pull()` come from).

:::

Everything looks good! Having made sure that the `ts` object contains the
individuals we want, let's move to the first exercise.

## Exercise 1: Computing nucleotide diversity

Our model of ancient human history makes a fairly clear prediction of what
should be the nucleotide diversity expected in the simulated populations.
**Compute the nucleotide diversity in all populations using the _slendr_
function
[`ts_diversity()`](https://www.slendr.net/reference/ts_diversity.html#ref-examples)
in your tree sequence `ts`. Add that command after your simulation code in your
new script in `popgen-stats.R` (do this even in the following exercises, to
keep everything in one place). Do you get numbers that (relatively between all
populations) match what would expect from the model given the $N_e$ that you
programmed for each of the lineages?** Check the figure created by the
`plot_model(model)` command earlier to get a refresher.

::: {.callout-note collapse="true" icon=false}
### Click to see the solution

**Population-based nucleotide diversity:**

Let's first get a named list of individuals in each group we want to be working
with (slendr tree-sequence statistic functions generally operate with this kind
of structure):

```{r}
sample_sets <- ts_names(ts, split = "pop")
sample_sets
```

We can use such `sample_sets` object to compute nucleotide diversity (pi)\ in
each population, in a bit of a similar manner to how we would do it with the
[standard _tskit_ in
Python](https://tskit.dev/tskit/docs/stable/python-api.html#tskit.TreeSequence.diversity):

```{r}
pi_pop <- ts_diversity(ts, sample_sets = sample_sets)
arrange(pi_pop, diversity)
```

You can see that this simple computation fits  the extreme differences in
long-term $N_e$ encoded by your _slendr_ demografr model, visualized here for
reference:

```{r}
plot_model(model)
```

:::

**Hint:** Nearly every _slendr_ statistic function interfacing with _tskit_
accepts a `ts` tree-sequence object as its first argument, with further arguments
being either:

- a vector of individual names representing a group of samples to compute a
statistic on,

- a (named) list of such vectors (each element of that list for a group of
samples with that name).

Although you can get all the above information by processing the table produced
by the `ts_samples()` function, _slendr_ provides a useful helper function
`ts_names()` which only returns the names of individuals as a vector 
(or a named list of such vectors, one vector per population as shown below).

When you call it directly, you get a plain vector of individual names:

```{r}
ts_names(ts)
```

This is not super helpful, unless we want to compute some statistic for _everyone_
in the tree sequence, regardless of their population assignment. A bit
more useful is to call the function like this, because it will produce a result
which can be immediately used as the `sample_sets =` argument mentioned in the
**Hint** above:

```{r}
ts_names(ts, split = "pop")
```

As you can see, this command returned a normal R list, with each element
containing a vector of individual names in a population. This means that we can
use standard R named-list indexing to get subsets of individuals like this:

```{r}
# this saves the named list of individual samples into a variable `names`
names <- ts_names(ts, split = "pop")

# this is how we can then access the names of all recorded Neanderthals
names["NEA"]

# this is how we can access the lists of names of two populations
names[c("EUR", "NEA")]
```

Many of the following exercises will use these kinds of tricks to instruct
various _slendr_ / _tskit_ functions to compute statistics on subsets of all
individuals sub-sampled in this way, so you can refer back to this **Hint**
later whenever you need help to compute some statistic.


---

**After you've computed nucleotide diversity per-population, compute it for
each individual separately (so, not for whole populations across all
individuals) using the same function `ts_diversity()`** (which will in this
setting give you the heterozygosity for each individual). **If you are familiar
with plotting in R, visualize the individual-based heterozygosities across all
populations.**

**Hint:** Following the previous hint on specifying individual names, you can
do this by giving the `ts_diversity()` function a vector of names as
`sample_sets =` instead of an R list of vectors like you did in the previous
exercise. One way to make plotting of the result easy is to first get a table
of all individuals using `pi_df <- ts_samples(ts)` and then add
heterozygosities in each of those individuals as a new column.

::: {.callout-note collapse="true" icon=false}
### Click to see the solution

**Per-individual heterozygosity:**

We can do this by passing the vector of individual names directory as the `sample_sets =` argument, rather than in a list of groups as we did above.

For convenience, we first get a table of all individuals (which of course
contains also their names) and in the next step, we'll just add their
heterozygosities as a new column:

```{r}
# get the table of all sample individuals
pi_df <- ts_samples(ts)
pi_df

# this is how we get names of all these individuals
pi_df$name

# and we can create a new `diversity` column of this table like this
pi_df$diversity <- ts_diversity(ts, sample_sets = pi_df$name)$diversity
pi_df
```

Let's plot the results using the ggplot2 package (because a picture is worth
a thousand numbers!)

```{r}
library(ggplot2)

ggplot(pi_df, aes(pop, diversity, color = pop, group = pop)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  theme_bw()
```

:::

## Exercise 2: Computing pairwise divergence

**Use the function
[`ts_divergence()`](https://www.slendr.net/reference/ts_divergence.html#ref-examples)
to compute genetic divergence between all pairs of populations. Similarly to
the previous exercise, do you get results compatible with our demographic model
in terms of the expectations given by the split times between the populations
in the model? Is there any indication of Neanderthal introgression in our
results (a bit of a spoiler for later)? What is missing to be able to
confidently say something about this in this case?**

**Hint:** Again, you can use the same concept of `sample_sets =` we discussed
in the previous exercise. In this case, the function computes _pairwise_
divergence between each element of the list given as `sample_sets =` (i.e.,
for each vector of individual names in that list).

::: {.callout-note collapse="true" icon=false}
### Click to see the solution

```{r}
# again, we can first extract the list of names of individuals in each population
names_list <- ts_names(ts, split = "pop")

# then we use this list in `ts_divergence()` to compute pairwise divergence values
div_df <- ts_divergence(ts, sample_sets = names_list)
arrange(div_df, divergence)
```

We can see that the pairwise nucleotide divergences between populations
recapitulate the known population/species relationships we would expect from
our model.

We can also see that the EUR-NEA divergence is "slightly lower" than the
AFR-NEA divergence, which could be due to NEA-EUR introgression, but we lack a
means of doing statistical significance test in this simple model. This is not
a big problem in a simulation because here we can more about investigating
trends in our statistics of interests rather than about formal hypothesis
testing, but something like jackknifing would be needed in a hypothetical
empirical study.

:::

## Exercise 3: Outgroup $f_3$ statistic

**Use the function `ts_f3()` to compute the outgroup $f_3$ statistic between
African-European, African-Neanderthal, and European-Neanderthal pairs versus a
Chimpanzee outgroup. For instance, the first statistic can be computed as
`ts_f3(ts, B = "AFR_1", C = "EUR_1", A = "CHIMP_1")`.**

::: aside

**Note:** The $f_3$ statistic is traditionally expressed as $f_3(A, B; C)$,
where C represents the outgroup. Unfortunately, in _tskit_ the outgroup is
named A, with B and C being the pair of samples from which we trace the length
of branches towards the outgroup, so the statistic is interpreted as $f_3(B, C;
A).$

:::

::: aside

**Note:** As a reminder, the higher the $f_3$ value, the more evolutionary
time has been shared between samples B and C.

:::

**How do the outgroup $f_3$ results compare to what you would expectat given
the phylogenetic relationships between populations in your model?**

**Do you see any impact of introgression on the $f_3$ value when a Neanderthal
is included in the computation (again, more on that later)?**

::: {.callout-note collapse="true" icon=false}
### Click to see the solution

```{r}
# The standard f3 formalism is this:
#
#    f3(A, B; C) = E[(A - C) * (B - C)]
#
# But in tskit and `ts_f3()`, A is the outgroup, so we have to reorganize the
# population labels...

# We can compute f3 for individuals...
ts_f3(ts, B = "AFR_1", C = "EUR_1", A = "CHIMP_1")

# ... but also whole populations (or population samples)
ts_f3(ts, B = sample_sets["AFR"], C = sample_sets["EUR"], A = "CHIMP_1")

ts_f3(ts, B = sample_sets["AFR"], C = sample_sets["NEA"], A = "CHIMP_1")

ts_f3(ts, B = sample_sets["EUR"], C = sample_sets["NEA"], A = "CHIMP_1")
```

Again, it looks like the branch from the outgroup to the AFR-NEA split is a bit
longer than the EUR-NEA split, which could be due to introgression, but unless
we would do proper jackknifing, this can only be interpreted as a trend, not a
formal test. As briefly discussed above (and much more below), in an empirical
situation, we would get confidence intervals and Z-scores for these values.

:::


## Exercise 4: `mode = "branch"` vs `mode = "site"`

**Repeat the exercise 3 by setting `mode = "site"` in the `ts_f3()` function
calls** (this is actually the default behavior of all _slendr_ tree-sequence
based _tskit_ functions). This will switch the _tskit_ computation to using
mutation counts along each branch of the tree sequence, rather than using
branch length themselves. **Why might the branch-based computation be a bit
better if what we're after is investigating the expected values of statistics
under some model?**


::: {.callout-note collapse="true" icon=false}
### Click to see the solution

See [this
tutorial](https://tskit.dev/tutorials/no_mutations.html#genealogy-based-measures-are-less-noisy)
(and particularly the directly linked section) for explanation.

:::



## Exercise 5: Outgroup $f_3$ statistic as a linear combination of $f_2$ statistics

You might have learned that any complex $f$-statistic can be expressed as a
linear combination of multiple $f_2$ statistics (which represent simple branch
length separating two lineages). **Verify that this is the case by looking up
equation *(20b)* in [this amazing
paper](https://academic.oup.com/genetics/article/202/4/1485/5930214) and
computing an $f_3$ statistic for any arbitrary trio of individuals of your
choosing using such linear combination of $f_2$ statistics.**

**Hint:** Here's that equation in a schematic form

```
      f3(A, B; C) = f2(A, C) + f2(B, C) - f2(A, B) / 2
```

::: {.callout-note collapse="true" icon=false}
### Click to see the solution

```{r}
# standard f3
ts_f3(ts, B = "AFR_1", C = "AFR_2", A = "CHIMP_1")

# a "homemade" f3 statistic as a linear combination of f2 statistics
# f3(A, B; C) = f2(A, C) + f2(B, C) - f2(A, B) / 2
homemade_f3 <- (
  ts_f2(ts, A = "AFR_1", B = "CHIMP_1")$f2 +
  ts_f2(ts, A = "AFR_2", B = "CHIMP_1")$f2 -
  ts_f2(ts, A = "AFR_1", B = "AFR_2")$f2
) / 2

homemade_f3
```

You can see that the `ts_f3()` implementation gives basically the same value as
our "homemade" implementation based on a linear combination of $f_2$
statistics, just as predicted by the underlying math.

:::

## Exercise 6: Detecting Neanderthal admixture in Europeans

Let's now pretend its about 2008, we sequenced the first Neanderthal genome,
and we are working on a project that will [change human evolution research
forever](https://www.science.org/doi/10.1126/science.1188021). We also have the
genomes of a couple of people from Africa and Europe, which we want to use to
answer the most burning question of all evolutionary anthropology: _"Do some
people living today carry Neanderthal ancestry?"_

Perhaps you've heard earlier about an $f_4$ statistic (or its equivalent, the
$D$ statistic) which can be used as a test of "treeness". As a recap, for some
"quartet" of individuals or population samples (whose names can be giving as
characters such as "EUR_1" to the `ts_f4()` function, as you can see in the
help page of `?ts_f4`), they can be used as a hypothesis test of whether the
history of those samples is compatible with there not having been an
introgression.

**Compute the $f_4$ test of Neanderthal introgression in EUR individuals using
the _slendr_ function `ts_f4()` (make sure to set `mode = "branch"` argument to
minimize stochastic noise).**

::: aside

**Note:** By default, each _slendr_ / _tskit_ statistic function operates on
mutations, and this will switch them to use branch length (as you might know,
$f$-statistics are mathematically defined using branch lengths in trees and
`mode = "branch"` does exactly that).

:::

**Hint:** If you haven't learned this earlier, to solve this exercise, you want
to compute and compare the values of these two statistics:

1. $f_4$(\<some African\>, \<another African\>; \<Neanderthal\>, \<Chimp\>)

2. $f_4$(\<some African\>, \<a test European\>; \<Neanderthal\>, \<Chimp\>),

here `<individual>` can be the name of any individual recorded in your
tree sequence, such as names you saw as `name` column in the table returned
by `ts_samples(ts)` (i.e. `"NEA_1"` could be used as a "representative"
\<Neanderthal\> in those equations, similarly for `"CHIMP_1"` as the fourth
sample in the $f_4$ quarted representing the outgroup).

<!--
To simplify things a lot, we can understand the above equations as comparing the
counts of so-called BABA and ABBA allele patterns between the quarted of samples
specified in the statistics:

$$
f_4(<African>, <X>; <Neand.>, <Chimp>) = \frac{\#BABA - \#ABBA}{\#SNPs}
$$
-->

The first $f_4$ statistic above is not expected to give values "too different"
from 0 (even in case of Neanderthal introgression into Europeans) because we
don't expect two African individuals to differ "significantly" in terms of how
much alleles they share with a Neanderthal (because their ancestors never met
Neanderthals!). However, the other statistic should be "significantly negative"
if there Neanderthal introgression into Europeans happened some time throughout
their history.

**Is the second of those two statistics "much more negative" than the first,
as expected assuming introgression from Neanderthals into Europeans?**

**Why am I putting "significantly" and "much more negative" in quotes in the
previous sentence? What are we missing here for this being a true hypothesis
test as you might be accustomed to from computing $f$-statistics using a tool
such as ADMIXTOOLS?** (We hinted at this issue a couple of times earlier.)

::: {.callout-note collapse="true" icon=false}
### Click to see the solution

Compute the difference in the amount of allele sharing between two African
individuals and a Neanderthal:

```{r}
f4_null <- ts_f4(ts, W = "AFR_1", X = "AFR_2", Y = "NEA_1", Z = "CHIMP_1", mode = "branch")
f4_null
```

Compute the difference in the amount of allele sharing between an African
individual vs European individual and a Neanderthal:

```{r}
f4_alt <- ts_f4(ts, W = "AFR_1", X = "EUR_1", Y = "NEA_1", Z = "CHIMP_1", mode = "branch")
f4_alt
```

We can see that the second test resulted in an f4 value about ~20 times more
negative than the first test, indicating that a European in our test carries
"significantly more" Neanderthal alleles compared to the baseline expectation
of no introgression established by the comparison to an African ...

```{r}
abs(f4_alt$f4 / f4_null$f4)
```

... although this is not a real test of significance because we have no Z-score
or standard error which would give us something like a p-value for the
hypothesis test, this evaluation of statistical expectations and trends using
simulations is extremely useful in research, because it allows us to test our
ideas on ground-truth models. With empirical data, we rarely (if ever) know
exactly what happened. With simulations, we always know which process generated
the data and how that process could have influenced our statistics of interest.

:::






## Exercise 7: Detecting Neanderthal admixture in Europeans v2.0

The fact that we don't get something equivalent to a p-value in these kinds of
simulations is generally not a problem, because we're often interested in
establishing a trend of a statistic under various conditions, and understand
under which conditions and in which way its _expected value_ behaves in a
certain way. If statistical noise is a problem, we work around this by
computing a statistic on multiple simulation replicates or even increasing the
sample sizes.

::: {.aside}
**Note:** To see this in practice, you can check out a
[paper](https://www.pnas.org/doi/10.1073/pnas.1814338116#fig02) in which I used
this approach quite successfully on a related problem.
:::

On top of that, p-value of something like an $f$-statistic (whether it's
significantly different from zero) is also strongly affected by quality of
the data, sequencing errors, coverage, etc. (which can certainly be examined
using simulations!). However, these are aspects of modeling which are quite
orthogonal to the problem of investigating the expectations and trends of
statistics given some underlying evolutionary model, which is what we're after
in these exercises.

That said, even in perfect simulated data, what exactly does
_"significantly different from zero compared to some baseline expectation"_
mean can be blurred by noise with simple single-individual comparisons that we
did above. Let's increase the sample size a bit to see if a statistical
pattern expected in $f_4$ statistic from our Neanderthal introgression model
becomes more apparent.

**Compute the first $f_4$ statistic (the baseline expectation between a pair of
Africans) and the second $f_4$ statistic (comparing an African and a European),
but this time on _all recorded Africans_ and _all recorded Europeans_,
respectively. Plot the *distributions* of those two sets of statistics. This
should remove most of the evolutionary stochastic noise and make a statistical
trend stand out much more clearly.**

**Hint:** Whenever you need to compute something for many things in sequence,
looping is very useful. One way to do compute, say, an $f_4$ statistic over
many individuals is by using this kind of pattern using R's looping function
`lapply()`:

```{r}
#| eval: false

# Loop over vector of individual names (variable x) and apply a given ts_f4()
# expression on each individual (note the ts_f4(..., X = x, ...) in the code)
X <- c("ind_1", "ind_2", ...)

list_f4 <- lapply(
  X,
  function(x) ts_f4(ts, W = "AFR_1", X = x, Y = "NEA_1", Z = "CHIMP_1", mode = "branch")
)

# The above gives us a list of data frames, so we need to bind them all into a
# single table for easier interpretation and visualization
df_f4 <- do.call(rbind, list_f4)
```

::: {.callout-note collapse="true" icon=false}
### Click to see the solution

This gives us list of vectors of the names of all individuals in each
population...

```{r}
sample_sets <- ts_names(ts, split = "pop")
# ... which we can then access like this
sample_sets$AFR # all Africans
sample_sets$EUR # all Europeans
```

Let's compute the $f_4$ statistic for all Africans... 

```{r}
f4_afr_list <- lapply(
  sample_sets$AFR,
  function(x) ts_f4(ts, W = "AFR_1", X = x, Y = "NEA_1", Z = "CHIMP_1", mode = "branch")
)
# ... and Europeans
f4_eur_list <- lapply(
  sample_sets$EUR,
  function(x) ts_f4(ts, W = "AFR_1", X = x, Y = "NEA_1", Z = "CHIMP_1", mode = "branch")
)
```

Bind each list of data frames into a single data frame:

```{r}
f4_afr <- do.call(rbind, f4_afr_list)
f4_eur <- do.call(rbind, f4_eur_list)

# add population columns to each of the two results for easier plotting
f4_afr$pop <- "AFR"
f4_eur$pop <- "EUR"

# bind both tables together
f4_results <- rbind(f4_afr, f4_eur)
```

Now we can visualize the results:

```{r}
f4_results %>%
  ggplot(aes(pop, f4, color = pop)) +
  geom_boxplot() +
  geom_jitter() +
  geom_hline(yintercept = 0, linetype = 2) +
  ggtitle("f4(AFR, EUR; NEA, CHIMP)") +
  theme_bw()
```

We can see that the $f_4$ statistic test of Neanderthal introgression in
Europeans indeed does give a much more negative distribution of values compared
to the baseline expectation which compares two Africans to each other.

This is why simulations are so powerful --- we establish models which generate
data under known population history, and then we can investigate what would be
the values of our statistics of interest under those models (and perhaps guess
which models are then closest to the hidden truth behind our empirical data).

:::


## Exercise 8: Trajectory of Neanderthal ancestry in Europe over time

There used to be a lot of controversy about the question of whether or not did
Neanderthal ancestry proportion in Europeans decline or not over the past 40
thousand years (see Figure 1 in [this
paper](https://www.pnas.org/doi/full/10.1073/pnas.1814338116) and Figure 2 in
[this paper](https://www.nature.com/articles/nature17993)).

**Your simulated tree sequence contains a time-series of European individuals
over time. Use the _slendr_ function `ts_f4ratio()` to compute (and then plot,
if you're familiar with data visualization in R) the proportion of Neanderthal
ancestry in Europe over time, commonly designated `alpha` in the literature.**
Use $f_4$-ratio statistic of the following form:

```{r}
#| eval: false
ts_f4ratio(ts, X = <vector individual names>, A = "NEA_1", B = "NEA_2", C = "AFR_1", O = "CHIMP_1")
```

**If a Neanderthal introgression happened at 3% tens of thousands of years ago,
what would you expect the proportion to be in the present day? Lower? About the
same?**


::: {.callout-note collapse="true" icon=false}
### Click to see the solution

```{r}
# Extract table with names and times of sampled Europeans (ancient and present day)
eur_inds <- ts_samples(ts) %>% filter(pop == "EUR")
eur_inds

# Compute f4-ration statistic (this will take ~30s) -- note that we can provide
# a vector of names for the X sample set to the `ts_f4ratio()` function
nea_ancestry <- ts_f4ratio(ts, X = eur_inds$name, A = "NEA_1", B = "NEA_2", C = "AFR_1", O = "CHIMP_1")

# Add the age of each sample to the table of proportions
nea_ancestry$time <- eur_inds$time
nea_ancestry

nea_ancestry %>%
  ggplot(aes(time, alpha)) +
  geom_point() +
  geom_smooth(method = "lm", linetype = 2, color = "red", linewidth = 0.5) +
  xlim(40000, 0) +
  coord_cartesian(ylim = c(0, 0.1)) +
  labs(x = "time [years ago]", y = "Neanderthal ancestry proportion") +
  theme_bw() +
  ggtitle("Neanderthal ancestry proportion in Europeans over time")

# For good measure, let's test the significance of the decline using a linear model
summary(lm(alpha ~ time, data = nea_ancestry))
```

In conclusion, assuming neutral drift, we don't expect the Neanderthal
trajectory to change, even after tens of thousands of years.

:::

## Exercise 9 (<u>hard bonus challenge</u>): The ADMIXTURE controversy

In the discussion about ADMIXTURE, you have heard about very different models
giving exactly the same ADMIXTURE result (see Figure 2 in [this
paper](https://www.nature.com/articles/s41467-018-05257-7/figures/2)).

**Write a simulation script for each of the three model topologies in the
figure linked above. Then use the function `ts_vcf()` to save a VCF file with
genotypes from each of those models, convert those VCF files into the PLINK
file format, and run ADMIXTURE to confirm the results of that paper.** Of
course, the PLINK conversion and ADMIXTURE analysis is something you're going
to have to do on the command-line, using the skills you've learned earlier in
this course.

## Exercise 10 (<u>hard bonus challenge</u>): How many unique f4 quartets are there?

If you've learned about $f$-statistics before, you've probably seen various
symmetries in $f_4$ (but also other $f$-statistics) depending on the
arrangement of the "quartet". As a trivial example, an $f_3(A; B, C)$ and
$f_3(A; C, B)$ will give you exactly the same value, and the same thing applies
even to more complex $f$-statistics like $f_4$.

**Use simulations to compute how many unique $f_4$ values involving a single
quartet are there.**

**Hint:** Draw some trees to figure out why could that be true. Also, when
computing `ts_f4()`, set `mode = "branch"` to avoid the effect of statistical
noise due to mutations.

::: {.callout-note collapse="true" icon=false}
### Click to see the solution

```{r}
# # install a combinatorics R package
# install.packages("combinat")

library(combinat)

# These are the four samples we can create quartet combinations from:
quartet <- c("AFR_1", "EUR_1", "NEA_1", "CHIMP_1")

# This is how we can generate all permutations:
quartets <- permn(quartet)
quartets

# How many permutations there are in total?
#   4! = 4 * 3 * 2 * 1 = 24
factorial(4)

# We should therefore have 24 different quartet combinations of samples
length(quartets)

# Loop across all quartets, computing the corresponding f4 statistic (we want
# to do this using branch lengths, not mutations, as the mutation-based computation
# would involve statistical noise)
all_f4s <- lapply(quartets, function(q) ts_f4(ts, q[1], q[2], q[3], q[4], mode = "branch"))

# Bind the list of f4 results into a single data frame and inspect the results
all_f4s <- bind_rows(all_f4s) %>% arrange(abs(f4))
print(all_f4s, n = Inf)

# Narrow down the results to only unique f4 values
distinct(all_f4s, f4, .keep_all = TRUE)
distinct(all_f4s, abs(f4), .keep_all = TRUE)
```

:::

