# Algorithmic thinking

⚠️⚠️⚠️
**This chapter is a work-in-progress!**
⚠️⚠️⚠️

## Part N: Generalizing code as a function

You now have all the individual pieces to construct a vector of allele
frequency trajectories of a given length. **Write a function `simulate_trajectory()`,
which will accept the following three parameters and returns the trajectory
vector as its output.**

- `p_init`: initial allele frequency
- `generations`: number of generations to simulate
- `Ne`: effective population size

Then **simulate three trajectories with the same parameters (`p_init = 0.5`,
`generations = 1000`, `Ne = 1000`) and visualize them on a single plot using
the functions `plot()` and `lines()`**.

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
simulate_trajectory <- function(p_init, generations, Ne) {
  trajectory <- c(p_init)

  # for each generation...
  for (gen in seq_len(generations)) {
    # ... based on the allele frequency in the current generation...
    p_current <- trajectory[gen]
    # ... compute the frequency in the next generation...
    n_next <- rbinom(Ne, 1, p_current)
    p_next <- sum(n_next) / Ne
    # ... and save it in the trajectory vector
    trajectory[gen + 1] <- p_next
  }
  
  return(trajectory)
}

traj1 <- simulate_trajectory(p_init = 0.5, generations = 1000, Ne = 1000)
traj2 <- simulate_trajectory(p_init = 0.5, generations = 1000, Ne = 1000)
traj3 <- simulate_trajectory(p_init = 0.5, generations = 1000, Ne = 1000)

plot(traj1, type = "l", xlim = c(1, 1000), ylim = c(0, 1), xlab = "generations", ylab = "allele frequency")
lines(traj2)
lines(traj3)
```

:::

## Part N: Running simulations in a loop

You have now generalized your code so that it is much more flexible by accepting
parameters as function inputs. Let's **utilize this self-contained function to
run 50 simulations using the function `lapply` and then plot the results
in a `for` loop.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
n_reps <- 20

traj_list <- lapply(
  1:n_reps,
  function(i) simulate_trajectory(p_init = 0.5, generations = 1000, Ne = 1000)
)

# create an empty plot (we have to specify the boundaries and other parameters!)
plot(NULL, type = "l", xlim = c(1, 1000), ylim = c(0, 1), xlab = "generations", ylab = "allele frequency")

# then overlay all trajectories in an automated way
for (traj in traj_list) {
  lines(traj)
}
```

:::


