# Installation -- _tidyverse_

The chapters of the book focused on data science rely exclusively on the
ecosystem of R packages called [_tidyverse_](https://www.tidyverse.org), plus a
set of functionality available in any basic R installation without any
additional setup needed. All of these packages are so widely used that it's
very easy to install them on practically any machine and any operating system.

**Please don't hesitate to get in touch with me over email at `contact [snail
sign] bodkan.net` if you run into any problems with this setup procedure or
anything else related to this! If you're reading this in preparation for one of
my in-person workshops, you can also catch me whenever you see me at the
teaching venue.**

## Installation steps

**Note:** It will be great if you do this setup on your personal laptop.
Getting everything to run locally is often much easier than trying to
install things on shared clusters or other HPC environments.

1. **I assume you already have [R](https://www.r-project.org) and
[RStudio](https://posit.co/download/rstudio-desktop/) installed on your
computer (any version will do).**

2. **Open RStudio and copy this into the R console to install the
_tidyverse_ packages that we will actually be using in the workshop:**

```{r}
#| eval: false
install.packages(c("dplyr", "ggplot2", "readr", "forcats", "cowplot", "palmerpenguins"))
```

3. **Then, in the same R console, run the following installation command:**

```{r}
#| eval: false
install.packages(c("rnaturalearth", "sf"))
```

This installs two very useful R packages for working with geospatial data.

**Note:** If you get an error during the installation, please copy-and-paste
the entire output from this command and send it to me via email. Or just talk
to me in person! The point 2. should work without any issues, but point 3.
might require a bit more setup, particularly on Linux or Mac. Don't worry if
things don't work 100%! Just reach out to me.

---

**If you've made it this far, you're good to go!**
