# Installation -- _tidyverse_

The chapters of the workbook focused on data science rely exclusively on the
ecosystem of R packages called
[_tidyverse_](https://www.tidyverse.org) (plus a set of functionality available
in any basic R installation without any additional setup needed). All of
these packages are so mainstream that it's completely trivial to install
them on practically any machine and any operating system.

**Please don't hesitate to get in touch with me over email at
`contact [snail sign] bodkan.net` if you run into any problems with this
setup procedure or with anything else related to the workshop! You can also
catch me whenever you see me in the teaching venue.**

## Installation steps

::: {.aside}
**Note:** It will be great if you do this setup on your personal laptop.
Getting everything to run locally is often much easier than trying to
install things on shared clusters or other HPC environments.
:::

1. I assume you already have [R](https://www.r-project.org) and
[RStudio](https://posit.co/download/rstudio-desktop/) installed on your computer
(any version will do).

2. Open RStudio and copy this into the R console to install the
_tidyverse_ packages that we will actually be using in the workshop:

```{r}
#| eval: false
install.packages(c("dplyr", "ggplot2", "readr", "forcats", "cowplot"))
```

3. Then, in the same R console, run the following installation command:

```{r}
#| eval: false
install.packages(c("rnaturalearth", "sf"))
```

This installs two very useful R packages for working with geospatial data.
Depending on how much time we have, there will be a session on the basics of
visualization of spatial information using R. These two packages are all that
are necessary for that.

::: {.aside}
**Note:** If you get an error during the installation, please copy-and-paste the
entire output from this command and send it to me via email. Or just talk to
me in person! The point 2. should work without any issues, but point 3. might
require a bit more setup, particularly on Linux or Mac. Don't worry if things
don't work 100%! Just reach out to me.
:::

---

**If you've made it this far, you're good to go!**
