# Installation

This entire workshop is organized around the population genetic simulation R
package called [_slendr_](https://slendr.net) and an assortment of other
useful R packages for data science, all of which will be automatically
installed alongside _slendr_.

The primary goal for the course is to make you comfortable programming complete
simulation and modeling scripts from scratch, entirely on your own. Even if you've
never "really programmed" before, it's not a problem. By the end of this
workshop, you will all be programmers!

**Please don't hesitate to get in touch with me over email at
`contact [snail sign] bodkan.net` if you run into any problems with this
setup procedure or with anything else related to the workshop! You can also
catch me whenever you see me in the teaching venue.**


## Installation steps

::: {.aside}
**Note:** It will be great if you do this setup on your personal laptop.
Getting everything to run locally is often much easier than trying to
install things on shared clusters or other HPC environments.
:::

1. I assume you already have [R](https://www.r-project.org) and
[RStudio](https://posit.co/download/rstudio-desktop/) installed on your computer
(any version will do).

2. Open RStudio and type this into the R console to install the _slendr_ package:

```{r}
#| eval: false
install.packages("slendr")
```

::: {.aside}
**Note:** If you get an error during the installation, please copy-and-paste the
entire output from this command and send it to me via email.
:::

3. After you have _slendr_ successfuly installed, create a dedicated Python
environment which it internally uses for simulation and tree-sequence analysis
by typing this into the R console again:

```{r}
#| eval: false
slendr::setup_env(agree = TRUE)
```

If everything worked, you should get an optimistic message at the end of the
entire procedure saying:

```         
======================================================================
Python environment for slendr has been successfuly created, and the R
interface to msprime, tskit, and pyslim modules has been activated.

In future sessions, activate this environment by calling init_env().
=======================================================================
```

::: {.callout-warning collapse="true"}
#### Click here if running `setup_env()` fails

If the `setup_env()` installation procedure above fails, try the following:

1. Delete the failed installation attempt:

```{r}
#| eval: false
slendr::clear_env(force = TRUE, all = TRUE)
```

2. Restart R by clicking (in RStudio) on `Session -> Restart R`.

3. Try installing it again, this time using `pip` as a Python installation method
(the default is `conda` which unfortunately fails fairly often):

```{r}
#| eval: false
slendr::setup_env(agree = TRUE, pip = TRUE)
```

::: {.aside}
**Note:** If you still get an error during the installation, please copy-and-paste
the entire output from this command and send it to me via email. Or just grab
me whenever you see me in person!
:::

:::




4. Paste this little testing _slendr_ simulation script into the R console to make
sure that everything works correctly. Don't read too much into the meaning of the
code, understanding (and being able to program) all this and more will be the point
of our workshop.

::: {.aside}
**Note:** If you managed to successfully run the installation steps 1-3
above, this simulation should finish running without any problems. If it does not work,
then there's something strange going on. Please copy-and-paste the entire output this
script produces in your R console (including the error) and send it to me via email.
:::

```{r}
#| collapse: true
library(slendr)
init_env()

A <- population("popA", time = 8000, N = 1000)
B <- population("popB", time = 4000, N = 1000, parent = A)
C <- population("popC", time = 4000, N = 1000, parent = A)
D <- population("popD", time = 3000, N = 1000, parent = C)

gf <- list(
  gene_flow(from = B, to = D, start = 1000, end = 800, rate = 0.2),
  gene_flow(from = C, to = D, start = 1000, end = 800, rate = 0.1)
)

model <- compile_model(
  populations = list(A, B, C, D),
  gene_flow = gf, generation_time = 30
)

ts <- msprime(model, sequence_length = 1e6, recombination_rate = 1e-8)
ts
```

The outcome of this script should approximately match what you see above (minus some
statistical noise manifesting in your numbers in the summary being slightly different).


5. Installing a couple of bonus R packages

In order to be able to run my own solutions to some of the exercises (particularly
some more advanced bonus exercises which will be entirely optional, and we very
likely won't have time to go through during the short span of the worksop), a
couple of additional R packages might be useful.

You can install those like this:

```{r}
#| eval: false
install.packages(c("combinat", "cowplot"))
```

---

**If you've made it this far, you're good to go!**
