# Installation -- _slendr_

The chapters of this book focused on population genomics and simulations rely
pretty much exclusively on the R package
[_slendr_](https://bodkan.github.io/slendr), plus an assortment of other useful
R packages for data science (all of which will be automatically installed
alongside _slendr_). Below are instructions how to prepare everything that will
be needed before the start of our workshop.

**Please don't hesitate to get in touch with me over email at `contact [snail
sign] bodkan.net` if you run into any problems with this setup procedure or
anything else related to this! If you're reading this in preparation for one of
my in-person workshops, you can also catch me whenever you see me at the
teaching venue.**

## Installation steps

**Note:** It will be great if you do this setup on your personal laptop.
Getting everything to run locally is often much easier than trying to install
things on shared clusters or other HPC environments.

1. **I assume you already have [R](https://www.r-project.org) and
[RStudio](https://posit.co/download/rstudio-desktop/) installed on your
computer (any version will do).**

2. **Open RStudio and copy this into the R console to install the _slendr_
package:**

```{r}
#| eval: false
install.packages("slendr")
```

**Note:** If you get an error during the installation, please copy-and-paste
the entire output from this command and send it to me via email.

3. **Paste this little testing simulation script into the R console to make
sure that everything works correctly.** Don't read too much into the meaning of
the code, understanding (and being able to program) all this and more will be
the point of our workshop.

**Note:** If you managed to successfully get through the steps 1-2 above, this
simulation should finish running without any problems. If it does not work,
please copy-and-paste the entire output this script produces in your R console
(including the error) and send it to me via email.

```{r}
#| collapse: true
library(slendr)

init_env(uv = TRUE)

A <- population("popA", time = 8000, N = 1000)
B <- population("popB", time = 4000, N = 1000, parent = A)
C <- population("popC", time = 4000, N = 1000, parent = A)
D <- population("popD", time = 3000, N = 1000, parent = C)

gf <- list(
  gene_flow(from = B, to = D, start = 1000, end = 800, proportion = 0.2),
  gene_flow(from = C, to = D, start = 1000, end = 800, proportion = 0.1)
)

model <- compile_model(
  populations = list(A, B, C, D),
  gene_flow = gf, generation_time = 30
)

ts <- msprime(model, sequence_length = 1e6, recombination_rate = 1e-8)
ts
```

The outcome of this script should approximately match what you see above (minus
some statistical noise manifesting in your numbers in the summary being
slightly different).

4. **Installing a couple of bonus R packages**

In order to be able to run my solutions to some of the exercises (particularly
some more advanced optional bonus exercises), a couple of additional R packages
might be useful. Feel free to only install those if you run into them at a
later point.

You can install those packages with this command:

```{r}
#| eval: false
install.packages(c("combinat", "cowplot", "smartsnp", "viridis", "ggrepel"))
```

---

**If you've made it this far, you're good to go!**
