# Installation -- _slendr_

The chapters of this book focused on population genomics and simulations rely
exclusive on the R package called [_slendr_](https://slendr.net) (plus an
assortment of other useful R packages for data science, all of which will be
automatically installed alongside _slendr_).

**Please don't hesitate to get in touch with me over email at `contact [snail
sign] bodkan.net` if you run into any problems with this setup procedure or
with anything else related to the workshop! If you're reading this in
preparation for one of my teaching workshop, you can also catch me whenever you
see me at the teaching venue.**

## Installation steps

::: {.aside}
**Note:** It will be great if you do this setup on your personal laptop.
Getting everything to run locally is often much easier than trying to
install things on shared clusters or other HPC environments.
:::

1. I assume you already have [R](https://www.r-project.org) and
[RStudio](https://posit.co/download/rstudio-desktop/) installed on your
computer (any version will do).

2. Open RStudio and copy this into the R console to install the _slendr_
package:

```{r}
#| eval: false
install.packages("slendr")
```

::: {.aside}

**Note:** If you get an error during the installation, please copy-and-paste
the entire output from this command and send it to me via email.

:::

3. Paste this little testing simulation script into the R console to make sure
that everything works correctly. Don't read too much into the meaning of the
code, understanding (and being able to program) all this and more will be the
point of our workshop.

::: {.aside}

**Note:** If you managed to successfully run the steps 1-3 above, this
simulation should finish running without any problems. If it does not work,
then there's something strange going on. Please copy-and-paste the entire
output this script produces in your R console (including the error) and send it
to me via email.

:::

```{r}
#| collapse: true
library(slendr)

init_env(uv = TRUE)

A <- population("popA", time = 8000, N = 1000)
B <- population("popB", time = 4000, N = 1000, parent = A)
C <- population("popC", time = 4000, N = 1000, parent = A)
D <- population("popD", time = 3000, N = 1000, parent = C)

gf <- list(
  gene_flow(from = B, to = D, start = 1000, end = 800, proportion = 0.2),
  gene_flow(from = C, to = D, start = 1000, end = 800, proportion = 0.1)
)

model <- compile_model(
  populations = list(A, B, C, D),
  gene_flow = gf, generation_time = 30
)

ts <- msprime(model, sequence_length = 1e6, recombination_rate = 1e-8)
ts
```

The outcome of this script should approximately match what you see above (minus
some statistical noise manifesting in your numbers in the summary being
slightly different).

4. Installing a couple of bonus R packages

In order to be able to run my solutions to some of the exercises (particularly
some more advanced optional bonus exercises), a couple of additional R packages
might be useful. Feel free to only install those if you run into them at a
later point.

You can install those packages with this command:

```{r}
#| eval: false
install.packages(c("combinat", "cowplot", "smartsnp", "viridis", "ggrepel"))
```

---

**If you've made it this far, you're good to go!**
