# _slendr_ models and "direction of time"

```{r}
library(slendr)
init_env()
```

## Encoding demographic models

### Forward direction of time

**Encode the demographic model in the following diagram using the basic _slendr_
components (`population()`, `gene_flow()`, `compile_model()`) and then compile
it using the function `compile_model()`, storing it in the `model` variable.
Then use the function `plot_model()` to visualize the model as we have
encoded, verifying it visually.**

**Note:** The time units are specified in the forward direction, so you will
need to specify the length of the simulation using the `simulation_length`
argument of `compile_model()`.

```{r}
library(slendr)
init_env()

p1 <- population("p1", time = 1, N = 1000)
p2 <- population("p2", time = 1000, N = 1000, parent = p1)
p3 <- population("p3", time = 2000, N = 1000, parent = p2)
p4 <- population("p4", time = 3000, N = 1000, parent = p3)

gf <- gene_flow(from = p2, to = p3, start = 3200, end = 3500, proportion = 0.1)

model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1, simulation_length = 4000
)

plot_model(model, proportions = TRUE)
```

```{r}
#| echo: false
fmodel <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1, simulation_length = 4000
)
```


---

**What happens when you don't specify the `simulation_length` argument of
the `compile_model()` function? Do you get an error if you leave out this
argument? Why?**

```{r}
#| error: true
model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1
)
```

Notice that we did not yet run any (potentially computationally costly)
simulation and _slendr_ gave us an informative error message as soon as it
encountered a potential error in our code. This is a very strong argument
for building our _slendr_ models step by step, executing individual commands
one after another, "conversing" with the R console in a way which lets _slendr_
catch our mistakes as early as possible.

This is opposed to the other way of working, in which we would first write an
entire (potentially very long) script and having trouble identifying where
and how we made a mistake.

---

### Backward direction of time

**As a practice for how does _slendr_ allow flexibility with direction of time
specification, try to convert your model into using "backward" direction of
time (i.e., the model will implicitly end at time 0 --- "the present day",
and start at the appropriate "time ago"). Again, visualize the final `model`
using the `plot_model()` function, which should get you (visually) exactly
the same plot.**

**In the previous exercise, leaving out the `simulation_length` argument
resulted in an error. What happens when you leave it out this time?**

**What happens when you set it to a number that is shorter than the indended
duration of the model in the schematic at the beginning, like setting
`simulation_length = 1200`? What would this imply about our programmed model?**

```{r}
library(slendr)
init_env()

p1 <- population("p1", time = 4000, N = 1000)
p2 <- population("p2", time = 3000, N = 1000, parent = p1)
p3 <- population("p3", time = 2000, N = 1000, parent = p2)
p4 <- population("p4", time = 1000, N = 1000, parent = p3)

gf <- gene_flow(from = p2, to = p3, start = 800, end = 500, proportion = 0.1)

model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1
)

plot_model(model, proportions = TRUE)
```

Indeed, both forward and backward models give the same figure (except for the
"direction" of time units along the y-axis):

```{r}
#| echo: false
bmodel <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1
)

cowplot::plot_grid(plot_model(fmodel), plot_model(bmodel))
```

As we can see in our solution, leaving out `simulation_length = 4000` did not
give an error this time because models encoded in "backward time units"
naturally end at time 0 (i.e., "the present day"). This means that setting
this argument works, but is redundant:

```{r}
model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1, simulation_length = 4000
)
```

On the other hand, if we do set it to a smaller number, we effectively
"truncate" the duration of the model. This is relatively rare, but it could be
useful in some circumstances, so it's worth mentioning here:

```{r}
model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1, simulation_length = 1200
)

plot_model(model)
```



#### Exercise: Creating a population assuming a forward direction of time

Let's start with the simplest possible _slendr_ simulation possible—a simulation
of the demographic history of a single population.

First, **create an R variable `pop` containing a _slendr_ population (using the
`population()` function) with the `name` "pop", at `time` 1 and a size `N` of
5000**.

**Note:** At this moment, the `time` value has a completely arbitrary unit to be
determined later. Additionally, unless otherwise specified, the size `N` will
always indicate the "effective population size" $N_e$.

**Execute your code which stores the population in the `pop` variable, then type
out `pop` in your R console. What does the summary report back to you?**

```{r}
pop <- population("pop", time = 1, N = 5000)
```

```{r}
print(pop)
```

#### Exercise: Compiling the model

```{r}
model <- compile_model(
  populations = pop,
  generation_time = 1,
  direction = "forward",
  simulation_length = 3000
)
```

```{r}
print(model)
```

```{r}
plot_model(model)
```

#### Exercise: Specifying generation time and time units

```{r}
model <- compile_model(
  populations = pop,
  generation_time = 30,
  direction = "forward",
  simulation_length = 1000
)
```

```{r}
print(model)
```

```{r}
plot_model(model)
```

#### Exercise: Backward direction of time

It's very easy to undestand models specified in time units going forward. However, in many practical situation in population genomics (and especially whenever we deal with ancient DNA samples), times of events specified in units such as "years before the present" are much easier to program, read, and interpret.

For instance, if we want to simulate a split of a lineage of anatomically modern humans from their chimpanzee relatives, it's much easier to say that it probably happened around six million years ago looking "backwards in time" and assuming that "present day" is counted at time 0, rather than having to recalculate all of these times now only into generations but also into generations going "forward" starting from a hypothetical time of 1 sometime in the past (like we did in the very first simulation above).

```{r}
pop <- population("pop", time = 1000, N = 5000)
```

```{r}
print(pop)
```

```{r}
model <- compile_model(
  populations = pop,
  generation_time = 1,
  direction = "backward"
)
```

```{r}
print(model)
```

```{r}
plot_model(model)
```

### Exercise: Resize events

#### Immediate resizes

```{r}
pop <- population("pop", time = 1, N = 5000) %>% resize(N = 2000, time = 500, how = "step")
```

```{r}
model <- compile_model(populations = pop, generation_time = 1, direction = "forward", simulation_length = 1000)
plot_model(model)
```

#### Exponential resizes

```{r}
pop <- population("pop", time = 1, N = 5000) %>% resize(N = 15000, time = 400, end = 600, how = "exponential")
```

```{r}
model <- compile_model(populations = pop, generation_time = 1, direction = "forward", simulation_length = 1000)
plot_model(model)
```

### Exercise: Encoding gene flow (admixture, introgression)

Of course, real-world populations don't just split from one another or change their size. A very common demographic process is gene flow (also called admixture or introgression in various context and different species).

**Use the function `gene_flow()` to program admixture from population "p1" into population "p2" at starting at `time` 100, with an `end` at time `200`. Set the total `proportion` of "p1" ancestry in "p2" at 0.2** (i.e., 20% of the final "p2" population will be derived from "p1").

```{r}
p1 <- population("p1", time = 1, N = 5000) %>% resize(N = 15000, time = 400, end = 600, how = "exponential")
p2 <- population("p2", time = 1, N = 5000) %>% resize(N = 15000, time = 400, end = 600, how = "exponential")
```

```{r}
model <- compile_model(populations = pop, generation_time = 1, direction = "forward", simulation_length = 1000)
plot_model(model)
```

**When you look at `?gene_flow` help page, you see that the function has a `rate` argument too. What does it do? Is there any way to get the same result (i.e., to encode the same gene-flow event) as you did just above, but using `rate` instead of `proportion`? When would you expect to use one or the other?**

### Exercise: Serialized models on disk

A very important part of _slendr_'s approach to modeling is to make simple things be trivially easy. This is largely why expressing even relatively complex models with _slendr_ requires so little code compared to the amount of programming needed with traditional powerful simulators such as [_msprime_](https://tskit.dev/msprime/) and [SLiM](https://messerlab.org/slim/), something you'll see (and hopefully appreciate!) when we get to programming larger models later.

Still, sometimes it's useful to peek behind the curtain and see what exactly is going on inside _slendr_ and how it operates.

**Type out the `model` object in the R console, paying attention to the description of the path. Use your operating system disk browsing program (Finder on macOS, File Explorer on Windows, or any other means of browsing files on other systems, such as the terminal shell) to see what files are hiding in that path.**

```{r}
list.files(model$path, full.names = TRUE)
```

```{r}
print_file <- function(model, file) cat(readLines(file.path(model$path, file)))
```

```{r}
print_file(model, "populations.tsv")
```

```{r}
print_file(model, "time_units.txt")
```

```{r}
print_file(model, "direction.txt")
```

```{r}
print_file(model, "description.txt")
```


