# Programming _slendr_ demographic models

Remember that every _slendr_ script for every exercise will begin with these two
lines of code:

```{r}
library(slendr)
init_env()
```

**Note:** For the time being, let's assume that each number indicating a time or
duration of some demographic event automatically implies times expressed in
units of generations, specified by integer numbers. We will discuss units of
time in _slendr_ in a later exercise.

### Exercise: Forward direction of time

**Encode the demographic model in the following diagram using the basic _slendr_
components (`population()`, `gene_flow()`, `compile_model()`) and then compile
it using the function `compile_model()`, storing it in the `model` variable.
Each time you write a line with a _slendr_ R command, execute it in your R
console by pressing Ctrl (or Cmd)+Enter, and then type out the name of the
variable with a given component (a population object, or the `model` model) in
the console and inspect the summary. When you finished writing the entire model,
use the function `plot_model()` to visualize the model as we have encoded,
verifying it visually.**

**Note:** The time units are specified in the forward direction, so you will
need to specify the length of the simulation using the `simulation_length`
argument of `compile_model()`.

**Note:** At this moment, the `time` value has a completely arbitrary unit to be
determined later. Additionally, unless otherwise specified, the size `N` will
always indicate the "effective population size" $N_e$.

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

As always, we start our new script by loading _slendr_ itself and activating
its environment:

```{r}
library(slendr)
init_env()
```

We begin by creating the first population, called p1, storing it in a variable
with the same name. When we execute this line and type `p1` in the R console,
we get an informative summary. Especially when building very complex models
with _slendr_, it pays of to build them like this, one command at a time, always
making sure that what we are programming matches what we want to program!

```{r}
p1 <- population("p1", time = 1, N = 1000)

p1
```

Here is, then, the complete solution:

```{r}
library(slendr)
init_env()

p1 <- population("p1", time = 1, N = 1000)
p2 <- population("p2", time = 100, N = 1000, parent = p1)
p3 <- population("p3", time = 200, N = 1000, parent = p2)
p4 <- population("p4", time = 300, N = 1000, parent = p3)

gf <- gene_flow(from = p2, to = p3, start = 320, end = 350, proportion = 0.1)

model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1, simulation_length = 400
)

plot_model(model, proportions = TRUE)
```

```{r}
#| echo: false
fmodel <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1, simulation_length = 400
)
```

:::

---

**What happens when you don't specify the `simulation_length` argument of
the `compile_model()` function? Do you get an error if you leave out this
argument? Why?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
#| error: true
model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1
)
```

Notice that we did not yet run any (potentially computationally costly)
simulation and _slendr_ gave us an informative error message as soon as it
encountered a potential error in our code. This is a very strong argument
for building our _slendr_ models step by step, executing individual commands
one after another, "conversing" with the R console in a way which lets _slendr_
catch our mistakes as early as possible.

This is opposed to the other way of working, in which we would first write an
entire (potentially very long) script and having trouble identifying where
and how we made a mistake.

:::

---

### Exercise: Backward direction of time

Unlike most other simulation frameworks, _slendr_ gives us a lot of
flexibility in how to encode the time units of our models, or even the
direction of time used by our models.

**To see how this works, try to convert your model into using "backward"
direction of time (i.e., the model will implicitly end at time 0 --- "the
present day", and start at the appropriate "time ago"). Again, visualize the
final `model` using the `plot_model()` function, which should get you (visually)
exactly the same plot.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
library(slendr)
init_env()

p1 <- population("p1", time = 400, N = 1000)
p2 <- population("p2", time = 300, N = 1000, parent = p1)
p3 <- population("p3", time = 200, N = 1000, parent = p2)
p4 <- population("p4", time = 100, N = 1000, parent = p3)

gf <- gene_flow(from = p2, to = p3, start = 80, end = 50, proportion = 0.1)

model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1
)

plot_model(model, proportions = TRUE)
```

Indeed, when we compare both figures side by side, both forward and backward
models give the same figure (except for the "direction" of time units along the
y-axis):

```{r}
#| echo: false
bmodel <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1
)

cowplot::plot_grid(
  plot_model(fmodel) + ggplot2::ggtitle("Forward time direction"),
  plot_model(bmodel) + ggplot2::ggtitle("Backward time direction")
)
```

:::

---

**In the previous exercise, leaving out the `simulation_length` argument
resulted in an error. What happens when you leave it out this time?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

As we can see in our solution, leaving out `simulation_length = 400` did not
give an error this time because models encoded in "backward time units"
naturally end at time 0 (i.e., "the present day"). This means that setting
this argument works, but is redundant:

```{r}
model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1, simulation_length = 400
)
```

:::

---

**What happens when you set this argument to a number that is shorter than the
intended duration of the model in the schematic at the beginning, like setting
`simulation_length = 1200`? What would this imply about our programmed model?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

On the other hand, if we do set it to a smaller number, we effectively
"truncate" the duration of the model. This is relatively rare, but it could be
useful in some circumstances, so it's worth mentioning here:

```{r}
model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1, simulation_length = 120
)

plot_model(model)
```

:::

---

### Exercise: Understanding common _slendr_ errors

As noted above, _slendr_ is designed around interactive use, executing little
bits of code at a time, making sure that potential errors are caught as soon
as we make them, rather than after spending lots of time programming long
simulation scripts.

However, software errors often confuse new programmers, especially because most
errors represent completely trivial typos and unintended mistakes rather than
conceptual issues with our model's logic. This is why it's a good idea to
spend a few moments to study the most common errors recognize why they occur in
isolated, controlled settings.

--

First, let's do a bit of a setup using things you already know.

**Create an R variable `pop` containing a _slendr_ population (using the
`population()` function) with the name "pA", at time 1 and with the $N_e$ of
5000. Then, create a population "pB" of the same size, which splits from "pA" at
time 100.**.

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
pA <- population("pA", time = 1, N = 5000)
pB <- population("pB", time = 100, N = 5000, parent = pA)
```

:::

When you do this, _slendr_ automatically infers that you intend to specify
time in your model flowing in the forward direction (i.e., times of events
move from smaller numbers to larger numbers). Not only that, from this point
on, _slendr_ also enforces all following demographic events to follow this
time direction!

**To test how this works, try to create a third population named "pC", which
splits from "pB" at time 80. What does _slendr_ do when you do this? What's the
closest appropriate value of pC's split time which would be consistent with
_slendr_'s assumptions for this example model?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

We get an error because in the context of the model so far (pA starting
at time 1 and pB splitting from it at time 100), pC would effectively split
at an earlier time than it's parent!

```{r}
#| error: true
pC <- population("pC", time = 80, N = 5000, parent = pB)
```

We can see that even setting the split time of pC to time 100 doesn't work,
because _slendr_ enforces that a population can only split from its parent
population at the time the latter already exists in the model:

```{r}
#| error: true
pC <- population("pC", time = 100, N = 5000, parent = pB)
```

In other words, the earliest that pC can split is time 101:

```{r}
pC <- population("pC", time = 101, N = 5000, parent = pB)
```

:::

---

**As another example, let's look at `gene_flow(), which programs gene flow
between two populations represented by two R variables. Let's say that we try to
program a 42% gene flow from `pA` to `pC` starting at time 200 and ending at
time 150, or a gene flow starting at time 150 and ending at time 200? Given the
assumptions that _slendr_ makes about the intended direction of time units in
our model, which of those two calls to `gene_flow()` will give error and why?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

Programming gene flow starting at time 200 and ending at time 150 effectively
implies a backward direction of time. This isn't a problem, necessarily,
except that the information about histories of pA and pC populations (actually,
only the pC population! can you tell why?) makes _slendr_'s `gene_flow()`
function enforce a forward direction of time.

```{r}
#| error: true
gf <- gene_flow(from = pA, to = pC, proportion = 0.42, start = 200, end = 150)
```

The second version works, of course, for the same reason:

```{r}
#| error: true
gf <- gene_flow(from = pA, to = pC, proportion = 0.42, start = 150, end = 200)
```

:::

---

**Note:** There are other cases in which even the enforcement of model
consistency described above still allows broken models to be programmed, which
is why `compile_model()` provides yet another set of sanity checks. But we will
leave those aside for now. If you end up running into these issues, this
exercise has already equipped you to understand why _slendr_ sometimes gives
errors and how to interpret them to fix your models.

### Exercise: Customization of time units

For more theoretical or general models, specifying time in generations going
forward (or backward) is generally enough. However, in more empirical settings,
encoding times in other units can be quite handy. For instance, imagine you work
with ancient DNA samples assigned with radiocarbon dates which use units of
"years before the present". Having to manually convert such times into
generations can get quite frustrating and cumbersome quickly, both when
programming models and when analyzing simulated sequence data.

This is where the `generation_time` argument of `compile_model()` comes
in---with this argument, we instruct _slendr_ on how should the various "genetic
time numbers" in our model code interpreted as "real-world numbers". Additionally,
there's another optional argument called `time_units` which we can use to
give the numbers an explicit description.

---

**Create a new script `slendr-models-human.qmd` script so that the time units
can be interpreted as "years before present (BP)", using the human generation
time of 30 years. Then visualize the model to make sure you've programmed
everything correctly.**

**Hint:** You will want to modify the times of all demographic events in your
model---originally in units of generations---so that they are effectively
30-times larger (so that the times represent human generation times), and make
them run from larger values to smaller values (so that they represent units
"years before present (BP)"). Finally, for clarity, you might want to use the
`time_units` argument of `compile_model()` to give the units this description.

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
library(slendr)
init_env()

p1 <- population("p1", time = 12000, N = 1000)
p2 <- population("p2", time = 9000, N = 1000, parent = p1)
p3 <- population("p3", time = 6000, N = 1000, parent = p2)
p4 <- population("p4", time = 3000, N = 1000, parent = p3)

gf <- gene_flow(from = p2, to = p3, start = 2400, end = 1500, proportion = 0.1)

model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 30,
  time_units = "years before present"
)

plot_model(model, proportions = TRUE)
```

Again, when we compare this model to the very first model (which used time units
of generations running forward in time), we see that they are completely
identical---only the presentation changes (giving us much more convenience), but
the population genetics theory underlying this model remains the same!

```{r}
cowplot::plot_grid(
  plot_model(model, proportions = TRUE) + ggplot2::ggtitle("'years before present"),
  plot_model(fmodel) + ggplot2::ggtitle("Forward time direction"),
  plot_model(bmodel) + ggplot2::ggtitle("Backward time direction"),
  nrow = 1
)
```

:::

---


### Exercise: Resize events

Of course, populations in traditional demographic models in evolutionary and
population genetics don't just split and mix---they also experience population
bottlenecks, population expansions, contractions, etc. The means to program
these demographic events is through the function `resize()`.

These events generally come in two variants, which we can call "step" resizes or
exponential resizes. 

#### Immediate resizes

```{r}
pop <- population("pop", time = 1, N = 5000) %>% resize(N = 2000, time = 500, how = "step")
```

```{r}
model <- compile_model(populations = pop, generation_time = 1, direction = "forward", simulation_length = 1000)
plot_model(model)
```

#### Exponential resizes

```{r}
pop <- population("pop", time = 1, N = 5000) %>% resize(N = 15000, time = 400, end = 600, how = "exponential")
```

```{r}
model <- compile_model(populations = pop, generation_time = 1, direction = "forward", simulation_length = 1000)
plot_model(model)
```

### Exercise: Encoding gene flow (admixture, introgression)

Of course, real-world populations don't just split from one another or change their size. A very common demographic process is gene flow (also called admixture or introgression in various context and different species).

**Use the function `gene_flow()` to program admixture from population "p1" into population "p2" at starting at `time` 100, with an `end` at time `200`. Set the total `proportion` of "p1" ancestry in "p2" at 0.2** (i.e., 20% of the final "p2" population will be derived from "p1").

```{r}
p1 <- population("p1", time = 1, N = 5000) %>% resize(N = 15000, time = 400, end = 600, how = "exponential")
p2 <- population("p2", time = 1, N = 5000) %>% resize(N = 15000, time = 400, end = 600, how = "exponential")
```

```{r}
model <- compile_model(populations = pop, generation_time = 1, direction = "forward", simulation_length = 1000)
plot_model(model)
```

**When you look at `?gene_flow` help page, you see that the function has a `rate` argument too. What does it do? Is there any way to get the same result (i.e., to encode the same gene-flow event) as you did just above, but using `rate` instead of `proportion`? When would you expect to use one or the other?**

### Optional exercise: Serialized models on disk

_Feel free to skip this exercise, unless you managed to run through everything
quickly and want to learn about _slendr_ internals!_

A very important part of _slendr_'s approach to modeling is to make simple things be trivially easy. This is largely why expressing even relatively complex models with _slendr_ requires so little code compared to the amount of programming needed with traditional powerful simulators such as [_msprime_](https://tskit.dev/msprime/) and [SLiM](https://messerlab.org/slim/), something you'll see (and hopefully appreciate!) when we get to programming larger models later.

Still, sometimes it's useful to peek behind the curtain and see what exactly is going on inside _slendr_ and how it operates.

**Type out the `model` object in the R console, paying attention to the description of the path. Use your operating system disk browsing program (Finder on macOS, File Explorer on Windows, or any other means of browsing files on other systems, such as the terminal shell) to see what files are hiding in that path.**

```{r}
list.files(model$path, full.names = TRUE)
```

```{r}
print_file <- function(model, file) cat(readLines(file.path(model$path, file)))
```

```{r}
print_file(model, "populations.tsv")
```

```{r}
print_file(model, "time_units.txt")
```

```{r}
print_file(model, "direction.txt")
```

```{r}
print_file(model, "description.txt")
```


