# Simulating data and encoding sampling events

### Exercise: Simulating genomic data

Once you have compiled a _slendr_ model and stored it in a variable, we can
finally simulate data from it. By default, a simulation from a _slendr_ model
always produces a [tree-sequence](https://tskit.dev/tutorials/what_is.html)
object as a result.

::: aside

**Note:** As discussed in our introduction to this session, tree sequences
provide an extremely efficient means to store and work with genomic data at a
massive scale. However, you can always get simulated data even in [traditional
file
formats](https://bodkan.net/slendr/reference/index.html#tree-sequence-format-conversion),
such as VCF, EIGENSTRAT, or a plain table of ancestral/derived genotype states.

In this section, we will be only working with tree sequences, because it's much
easier and faster to get interesting statistics from it directly in R, but you
can always compute the same statistics from simulated VCF or EIGENSTRAT files
using any of the available software methods as well.

:::

There are two simulation engines built into _slendr_ implemented by functions
`msprime()` and `slim()`. For traditional, non-spatial, neutral demographic
models, the `msprime()` engine is much more efficient, so we'll be using that
for the time being. However, from a population genetics theoretical perspective,
both simulation functions will give you the same results for any given compiled
_slendr_ model (up to some level of stochastic noise). The alternative engine
implemented by the `slim()` function plays an important role for spatial
simulations and for simulations of natural selection.

As we discussed in the introduction, here's how you can use simulate a tree
sequence from a compiled _slendr_ `model`:

```{r}
#| eval: false
#| code-fold: false
ts <- msprime(
  model,
  sequence_length = <length of the simulated [bp]>,
  recombination_rate = <uniform recombination rate [crossovers per bp per generation]>
)
```

You will be seeing this kind of pattern over and over again.

---

**Simulate a tree sequence from your compiled `model` using the `msprime()`
engine and store it in a variable `ts` as shown right above. Set the parameters
`sequence_length = 1e6` (length of the sequence in bp) and `recombination_rate
= 1e-8` (crossover events per base pair per generation).**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

This command simulates a _tskit_ tree sequence from a _slendr_ model. Again,
note that we don't have to write any _msprime_ or _tskit_ code to implement the
simulation because _slendr_ does all the work for us!

```{r}
ts <- msprime(model, sequence_length = 1e6, recombination_rate = 1e-8)
```

:::

Similarly to the symbol `model`, whenever we from now on refer to `ts`, we will
be talking about some simulated tree-sequence object stored in an R variable
`ts`.

## Exercise: Inspecting the tree sequence

As we will see throughout the entire rest of this course, _slendr_ makes it
possible to compute a wide array of popular population genetic statistics using
its R interface to the [Python module
_tskit_](https://tskit.dev/tskit/docs/stable/python-api.html).

**Type out the `ts` object in the terminal --- what do you see? Try to relate
some of the information in the presented summary to our overview discussion of
tree sequences.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

You should get a summary of a tree-sequence object with some lower-level
technical details about the genomic sequence which it encodes:

```{r}
ts
```

:::

---

In the `ts` summary, you probably noticed there are no mutations! As we will see
later, the magic of tree sequences allows us to compute nearly any conceivable
population genetic statistic even without mutations. That said, particularly for
exploring statistics computed by various third-party software outside of R,
having mutations can be often useful.

**Modify the part of your script which simulates the `ts` object, adding a `%>%
ts_mutate(mutation_rate = 1e-8)` after the `msprime(...)` function call. After
you execute this new command and type `ts` in the R console to inspect the
summary, you should see that mutations have been added.**

::: aside

**Note:** Even though mutations are generally not needed, there's almost never
any huge downside to adding them, so I recommend you always use both `msprime()`
and `ts_mutate()` in your models.

:::

```{r}
ts <- msprime(model, sequence_length = 1e6, recombination_rate = 1e-8) %>%
  ts_mutate(mutation_rate = 1e-8)
```

---

There are also two important _slendr_ functions called `ts_samples()` (which
retrieves the "symbolic names" and dates of all recorded individuals at the end
of a simulation) and `ts_nodes()`. **You can run them simply as
`ts_samples(ts)` and `ts_nodes(ts)`. How many individuals (i.e. diploid
samples) are in your tree sequence as you simulated it? How is the result of
`ts_nodes()` different from `ts_samples()`?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

We can see that our tree sequence contains a massive number of individuals. Too
many, in fact --- we recorded every single individual alive at the end of our
simulation, which is something we're unlikely to be ever lucky enough to have,
regardless of which species we study.

```{r}
#| message: false
ts_samples(ts)
ts_samples(ts) %>% nrow()

library(dplyr)
ts_samples(ts) %>% group_by(pop) %>% summarize(n = n())
```

The function `ts_nodes()` returns a table similar to the one produced by
`ts_table(ts, "nodes")` above, except that it contains additional _slendr_
metadata (names of individuals belonging to each node, spatial coordinates of
nodes for spatial models, etc.). This makes it a bit more useful for analyzing
tree-sequence data than the "low-level" functions, if that's something you ever
need to do.

```{r}
ts_nodes(ts) %>% head(5)
```
:::

## Exercise 5: Scheduling sampling events

In the table produced by `ts_samples()` in the previous exercise, you saw that
the tree sequence we simulated recorded "everyone" living at the end of the
simulation which amounted to a few tens of thousands of individuals. However,
unless we're very lucky, it's extremely unlikely that we'll ever have a
sequence of every single individual in a population that we study. To get a
little closer to the scale of the genomic data that we usually work with on a
day-to-day basis (and to avoid wasting computational resources on unnecessarily
massive simulations), we can restrict our simulation to only record a subset of
individuals.

We can precisely define which individuals (from which populations, and at which
times) should be recorded in a tree sequence using the _slendr_ function
`schedule_sampling()`. For instance, in our `model` of introgression, we have
created _slendr_ populations stored in variables `eur` and `afr`, and we can
include the recording of the genomes of five individuals from both of those
populations at times 10000 (years ago) and 0 (present-day) in our model with
the following code:

```{r}
#| eval: false
pop_schedule <- schedule_sampling(
  model,                     # the model to record genomes from
  times = c(10000, 0),       # times at which to record genomes
  list(eur, 5), list(afr, 5) # which populations and how many genomes to record
)
```

**If you type the variable `pop_schedule` in the R console, you will see that
this function returns a plain data frame (just try it).** As such, we can
create multiple of such schedules (of arbitrary complexity and granularity) and
then bind them together into a single sampling schedule with a single line of
code, like this (all this code might look overwhelming, but try to read it line
by line):

```{r}
# Note that the `times =` argument of the `schedule_sampling()` function can be
# a vector of times like here...
ancient_times <- c(40000, 30000, 20000, 10000)
eur_samples <- schedule_sampling(model, times = ancient_times, list(eur, 1))

# ... but also just a single number like here
afr_samples <- schedule_sampling(model, times = 0, list(afr, 1))
nea_samples <- schedule_sampling(model, time = 60000, list(nea, 1))

# But whatever the means you create the individual sampling schedules with, you
# can always bind them all to a single table with the `rbind()` function
schedule <- rbind(eur_samples, afr_samples, nea_samples)

# This is then the final "recording schedule":
schedule
```

---

**Using the function `schedule_sampling` and with the help of `rbind` as shown
in the previous code example, program the sampling of the following sample sets
at given times:**

| time  | population | \# individuals |
|-------|:-----------|----------------|
| 70000 | nea        | 1              |
| 40000 | nea        | 1              |
| 0     | chimp      | 1              |
| 0     | afr        | 5              |
| 0     | eur        | 10             |


**Then, schedule the sampling of a single `eur` individual at each of the
following times (i.e., every two thousand years between 40 kya and the
present-day):**

```{r}
t <- seq(40000, 2000, by = -2000)
```

**At the end, bind all individual schedules into a single variable `schedule`
using the function `rbind()`. In total, this should give you a data frame
containing the sampling schedule for the recording of of the genomes of 38
simulated individuals. Verify that this is true by summing the numbers in the 
`n` column of the final data frame `schedule`.**


::: aside
**Note:** You can provide a vector variable (such as `t` in this example) as the
`times =` argument of `schedule_sampling()`.
:::

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

First we scheduled the sampling of two Neanderthals at 70kya and 40kya:

```{r}
nea_samples <- schedule_sampling(model, times = c(70000, 40000), list(nea, 1))
nea_samples # (this function produces a plain old data frame!)
```

Then we schedule one Chimpanzee sample, 5 African samples, and 10 European samples:

```{r}
present_samples <- schedule_sampling(model, times = 0, list(chimp, 1), list(afr, 5), list(eur, 10))
```

Then we also schedule the recording of one European sample between 50kya and
2kya, every 2000 years:

```{r}
times <- seq(40000, 2000, by = -2000)
emh_samples <- schedule_sampling(model, times, list(eur, 1))
```

Because those functions produce nothing but a data frame, we can bind
individual sampling schedules together at the end:

```{r}
schedule <- rbind(nea_samples, present_samples, emh_samples)
schedule
```

Let's verify we scheduled the recording of genomes of everyone we needed:

```{r}
sum(schedule$n)
```

:::

**Then, verify the correctness of your overall sampling `schedule` by
visualizing it together with your `model` using the function `plot_model()`
(take a look at its manpage `?plot_model` to see how to do this).**

::: aside

**Note:** As you've seen above, the visualization is often a bit wonky and
convoluted with overlapping elements and it can be even worse with samples
added, but try to experiment with arguments to `plot_model` described above to
make the plot a bit more helpful for sanity checking.

:::

```{r}
#| eval: false
plot_model(model, samples = schedule)
```

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
plot_model(model, sizes = FALSE, samples = schedule)
```

```{r}
plot_model(model, sizes = FALSE, log = TRUE, samples = schedule)
```
:::

## Exercise 6: Simulating a defined set of individuals

We now have both a compiled _slendr_ `model` and a well-defined sampling
`schedule`, so let's finally simulate some data!

**Use your combined sampling `schedule` to run a new tree-sequence simulation
from your model (again using the `msprime()` function), this time restricted to
just those individuals scheduled for recording. You can do this by running the
following command (note that this also overlays mutations over our genealogies,
which is something we didn't do before).** This will give you a new
tree-sequence object stored in the `ts` variable.

```{r}
#| eval: false
ts <-
  msprime(
    model, sequence_length = 100e6, recombination_rate = 1e-8, 
    samples = schedule
  ) %>%
  ts_mutate(mutation_rate = 1e-8)
```

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

The command below will likely take a few minutes to run, so feel free to go
down from 100 Mb sequence_length to even 10Mb (it doesn't matter much).
(The `random_seed =` argument is there for reproducibility purposes.)

```{r}
#| eval: false
ts <-
  msprime(model, sequence_length = 100e6, recombination_rate = 1e-8, samples = schedule, random_seed = 1269258439) %>%
  ts_mutate(mutation_rate = 1e-8, random_seed = 1269258439)
# Time difference of 2.141642 mins
```


If you're bothered by how long the simulation takes, feel free to call these two lines
to 100% reproduce my results without any expensive computation:

```{r}
#| eval: false
url_path <- "https://raw.githubusercontent.com/bodkan/simgen/refs/heads/main/files/popgen/introgression.trees"
ts_path <- tempfile()
download.file(url_path, destfile = ts_path, mode = "wb")
ts <- ts_read(file = ts_path, model = model)
ts
```

```{r}
#| echo: false
#| eval: false
# We can save a tree sequence object using a slendr function `ts_write` (this
# can be useful if we want to save the results of a simulation for later use).
if (dir.exists(here::here("files/popgen"))) {
  ts_write(ts, here::here("files/popgen/introgression.trees"))
}
```

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| eval: false -->
<!-- if (dir.exists(here::here("files/popgen"))) { -->
<!--   model <- read_model(here::here("files/popgen/introgression")) -->
<!--   ts <- ts_read(file = here::here("files/popgen/introgression.trees"), model = model) -->
<!-- } -->
<!-- ``` -->

:::

---

**Verify that your simulated `ts` object really does contain only the
explicitly recorded samples, either by typing `ts` into the R console or,
better still, using the function `ts_samples()`.**

::: aside

**Note:** When you think about it, it's quite astonishing how fast _msprime_
and _tskit_ are when dealing with such a huge amount of sequence data from tens
of thousands of individuals on a normal laptop!

:::

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

Inspect the (_tskit_/Python-based) summary of the new tree sequence
(note the much smaller number of "sample nodes"!):

```{r}
ts
```

Get the table of all recorded samples in the tree sequence;

```{r}
ts_samples(ts)
```

We can also put some _tidyverse_ skills to use and compute the count of
individuals at different time points:

```{r}
library(dplyr)

ts_samples(ts) %>%
  mutate(age_group = if_else(time == 0, "present-day", "ancient")) %>%
  group_by(pop, age_group) %>%
  summarize(n = n()) %>%
  select(age_group, pop, n)
```

:::
