

The genius of the tree-sequence data structure rests on its elegant table-based
implementation (much more information on that is
[here](https://tskit.dev/tskit/docs/stable/data-model.html)). _slendr_ isn't
really designed to run complex low-level manipulations of tree-sequence data
(like iterating over individual trees, mutations on those trees, etc.) as its
strength lies in the convenient interface to population genetic statistical
functions implemented by _tskit_, but it does contain a couple of functions
which can be useful for inspecting the lower-level nature of a tree sequence.
Let's look at some of them right now.

**Use the `ts_table` function to inspect the low-level table-based
representation of a tree sequence. Take a look at `?ts_table` manual page to
learn how to use this function. How many nodes does it have, and how many
edges? Does the tree sequence contain any mutations? If not, why, and how can
we even do any population genetics with data without any mutations?** As you're
answering these questions, take a look at at the [following
figure](https://tskit.dev/tutorials/_images/693660726ad4851fe5c0a4d04264dc4e79a2f042cc657a10e46395f13caccb75.svg)
(this was made from a different tree sequence than you have, but that's
alright) to help you relate the information in the individual tables (node
table, edge table, etc.) to your tree sequence.

::: aside

**Note:** This exercise will not teach you much about how are simulations of
tree sequences helpful for your research, but it should convince you that the
final product of a _slendr_ simulation really is a data structure encoding the
full extent of a genomic variation in your sample (either simulated or
empirical). You don't have to study these tables in detail, this is just to
remind you that they are always there, under the hood.

:::

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

The _slendr_ helper function `ts_table()` accepts two arguments: the first one
is a tree-sequence object `ts`, the second indicates which of the tree-sequence
tables we want to inspect:

```{r}
ts_table(ts, "nodes")
ts_table(ts, "edges")
ts_table(ts, "individuals")
```

We didn't simulate any mutations, so we only have genealogies for now:

```{r}
ts_table(ts, "mutations")
ts_table(ts, "sites")
```

:::

---