
### Option 2: Star Wars characters

If you need a break from population genetics and biology and would still
like to practice your _tidyverse_ and _ggplot2_ skills, here's a fun data
set for you with various metadata about Star Wars characters, with a list
of questions you can try answering. Every single question can be answered
using the tools you've already encountered here!

```{r}
library(dplyr)

starwars
```

1. What is the distribution of counts of sexes of the characters?
2. What is the distribution of counts of genders of the characters?
3. Who is the oldest character?
4. Who is youngest?
5. What's the most frequently featured homeworld planet?
6. What is the weight of Jar Jar Binks?

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

This is way too many questions to provide solutions too with extensive comments.
Luckily, you should be able to guess which code answers which question in which
way, so just experiment!

```{r}
ggplot(starwars, aes(sex)) + geom_bar()
```

```{r}
ggplot(starwars, aes(gender)) + geom_bar()
```

```{r}
ggplot(starwars, aes(height)) + geom_histogram()
```

```{r}
ggplot(starwars, aes(mass)) + geom_histogram()
```

There's a clear massive (literally) outlier. Is this a data-entry error, or
could it possibly be [this guy](https://www.starwars.com/databank/jabba-the-hutt)?

```{r}
filter(starwars, mass > 1000)
```

Jabba ruins everything, so let's cap the mass scale so that we can see some
patterns:

```{r}
starwars %>%
filter(!is.na(gender)) %>%
ggplot(aes(gender, mass, fill = gender)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  coord_cartesian(ylim = c(0, 250))
```

```{r}
starwars %>%
filter(!is.na(sex)) %>%
ggplot(aes(sex, mass, fill = sex)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  coord_cartesian(ylim = c(0, 250))
```


```{r}
ggplot(starwars, aes(mass, height)) + geom_point()
```

```{r}
ggplot(starwars, aes(mass, height)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 250))
```

```{r}
ggplot(starwars, aes(mass, height)) +
  geom_point() +
  geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(0, 250))
```

```{r}
ggplot(starwars, aes(mass, height)) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlim(c(0, 250))
  # coord_cartesian(xlim = c(0, 250))
```

:::