# Manipulating tabular data

Before we begin, let's introduce the two stars of the _tidyverse_ ecosystem,
which we will be using here:

1. _dplyr_ is a centerpiece of the entire R data science environment, providing
important functions for data manipulation, data summarization, filtering, etc.
of tabular data;

2. _readr_ is an R package which provides very convenient functions for reading
(and writing) tabular data. Think of it as a set of better alternatives to
base R functions such as `read.table()`, etc.

Every single script you will be writing in this session will begin with these
two lines of code.

```{r}
library(dplyr)
library(readr)
```

Let's also introduce a second star in this session, our example data set.
This commands reads a metadata table from a recent huge aDNA paper on the
history or the Holocene in West Eurasia, dubbed "MesoNeo" 
([reference](https://www.nature.com/articles/s41586-023-06865-0)).
You can read it like this, which will save it to a variable `df`.
_Everything_ in the section on tidyverse will revolve around this data.

```{r}
df <- read_tsv("https://tinyurl.com/qwe-asd-zxc")
```


## Part XYZ: 

## Part XYZ: Filtering rows

**Base R indexing recap**

## Part XYZ: Selecting columns

## Part XYZ: Summarizing data