# Data visualization

In this chapter, we will be delving into the data visualization R package
called [_ggplot2_](https://ggplot2.tidyverse.org). Here is how the start of
our solutions script for the day will look like. Note the addition of
`library(ggplot2)` to the list of R packages used in the previous chapter.
Although you could use _ggplot2_ without the rest, it ties so neatly with
every other part of the _tidyverse_ ecosystem that you almost always end up
using them together.

```{r}
#| message: false
library(ggplot2) # the star of the day!
library(dplyr)
library(tidyr)
library(readr)

df <- read_tsv("https://tinyurl.com/simgen-ibd")
```

**Create a new R script in RStudio, (`File` `->` `New file` `->` `R Script`) and
save it somewhere on your computer as `tidy-viz.R` (`File` -> `Save`). Copy
the chunk of code above into it and let's get started!**

```{r}
#| eval: false
#| echo: false
# load("~/Desktop/neo.impute.1000g.ibd.filter.strict.RData")
# ibd <- dd[dd$chromosome == 21, ]
# ibd <- dplyr::as_tibble(ibd)
# ibd <- dplyr::select(ibd, sample1, sample2, chrom = chromosome, start = posStart, end = posEnd)
# readr::write_tsv(ibd, "files/tidy/ibd.tsv.gz")
```
