# Visualization with _ggplot2_

## Introduction

In this chapter, we will be delving into the data visualization R package
called [_ggplot2_](https://ggplot2.tidyverse.org), which is possibly the
most famous piece of the _tidyverse_ ecosystem. So much so that people who
otherwise don't use any _tidyverse_ functions (or even who don't even use
R for data analysis itself) still use _ggplot2_ for making figures. It really
is that good.

## Part 1: Visualizing patterns in our IBD data

**First, let's start from a clean slate, and leverage the modularized
pipeline we created for our IBD data at the very end of our second _tidyverse_
session. Create a new R script in RStudio
(`File` `->` `New file` `->` `R Script`), and save it somewhere on your
computer as `tidy-viz.R` (`File` `->` `Save`). Copy the following bit of code
which will execute your IBD and metadata processing pipeline into that script.**

The following exercises will focus on basics of visualization of various
types of statistical summaries, using your IBD data set as a representative
of typical data we deal with when doing data science.

```{r}
#| eval: false
library(ggplot2) # the star of the day!

library(dplyr)
library(readr)

# source our utility functions so that we have them available
source("ibd_utils.R")

# download and process the metadata and IBD data set
metadata <- process_metadata(bin_step = 10000)
ibd_segments <- process_ibd()

# combine the IBD table with metadata information
ibd_merged <- join_metadata(ibd_segments, metadata)
```

```{r}
#| echo: false
#| output: false
library(ggplot2) # the star of the day!

library(dplyr)
library(readr)

source(here::here("files/repro/ibd_utils.R"))

metadata <- process_metadata(bin_step = 10000)
ibd_segments <- process_ibd()

ibd_merged <- join_metadata(ibd_segments, metadata)
```

**Let's appreciate the modularization aspect of the work we did earlier.
Practically speaking, all of that work is now hidden behind three custom
functions: `process_metadata()`, `process_ibd()` and `join_metadata()`.
We can reuse this code anywhere and if we ever need to update something,
we just have to change it in a single place: the `ibd_utils.R` script.
This is a very important example of good programming practice.**

### Exercise 1: Concept of layers

An important (and even elegant) concept of building visualizations
using _ggplot2_ is the idea of "layering". We will begin introducing this idea
from the simplest possible kind of visualization, which is plotting
the counts of observations of a categorical (i.e. discrete) variable,
**but first, what actually happens when you call a function `ggplot()` on its own?
How about when you call `ggplot(metadata)`, i.e. when we tell `ggplot()` we want
to be plotting something from out `metadata` table?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

We get an empty "canvas"! Almost as if we were an artist starting to paint
something, to use a slightly pretentious metapho.

```{r}
ggplot()
```

Saying that we want to plot `metadata` doesn't actually change anything. We
get the same empty canvas, with _ggplot2_ basically waiting for us to
give it more instructions.

```{r}
ggplot(metadata)
```

:::

As mentioned in my solution, calling `ggplot()` on its own basically only
sets up an empty "canvas", as if we were about to start painting something.

This might sound like a silly metaphor, but when it comes to
_ggplot2_ plotting, this is actually remarkably accurate, because it is
exactly how  the above-mentioned idea of the "layering grammar" of _ggplot2_
works in practice.

There are two important components of every _ggplot2_ figure:

#### 1. **"Aesthetics"**

Aesthetics are a specification of the _mapping_ of columns of a data frame
to some kind of visual property of the plot we want to create (position
along the `x`, or `y` axes, `color`, `shape,` etc.). We specify these
mapping aesthetics with a function called `aes()` which you will see later
many times.

#### 2. **"Geoms"**

Geoms are graphical elements to be plotted (histograms, points,
lines, etc.). We specify those using functions such as `geom_bar()`, 
`geom_histogram()`, `geom_point()`, and many many others.

#### **How do "aesthetics" and "geoms" fit together?**

We can think about it this way:

- "geoms" specify what our data should be plotted as in terms of "geometrical
elements" (that's where the name "geom" comes from)
- "aesthetics" specify how those geoms should look like (so, the aesthetic
visuals features of geoms, literally)

**This is all very abstract, but don't worry. Refer to this as a guideline
when you're implementing the exercises in practice!**


### Exercise 2: Distribution of a categorical variable

Let's start very simply, and introduce the layering aspect of _ggplot2_
visualization package step by step. **Our data frame has a column `age_bin`,
which contains the assignment of each individual into a respective time
period. Let's start by visualizing the count of individuals in each time bin.**

**Run the following code in your R console:**

```{r}
#| eval: false
ggplot(metadata, aes(x = age_bin))
```

**It specifies the aesthetic 
mapping which will map our first variable of interest (column `age_bin` in
our data frame) to the `x`-axis of our eventual figure. What do you get
when you run it? How is the result different from just calling `ggplot(metadata)`
without anything else, like you did above?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

Let's take it step by step:

1. Again, saying what data set we want to plot doesn't do anything (_ggplot2_
doesn't know what to plot yet!):

```{r}
ggplot(metadata)
```

2. mapping an aesthetic still doesn't do anything, because we didn't instruct
_ggplot2_ to plot anything even now:

```{r}
ggplot(metadata, aes(x = age_bin))
```

A-ha! We see a difference! _ggplot2_ now _mapped_ the `x` axis against
values of the `age_bin` variable. Let's verify this by looking at all possible
values of that variable:

```{r}
table(metadata$age_bin)
```


But even then, the plot still actually doesn't show anything!
Remember, we only instruct _ggplot2_ to map a variable `age_bin` to an
`x`-axis of a figure... but we didn't say _what_ we want to visualize
(i.e. a "geom").

To make a visualization, both aesthetic mapping and some kind of geom is needed! We will do this in the next step.

:::


#### `geom_bar()`

In the previous exercise, you instructed `ggplot()` to work with the `metadata` table and you mapped
a variable of interest `age_bin` to the x-axis (and therefore set up a
"mapping aesthetic"). **To make a complete figure, you need to add a layer with
a "geom". In _ggplot2_, we add layers with the `+` operator**, mimicking the
"painter metaphor" of adding layers one by one, composing a more complex
figure with each additional layer.

**Add the `+ geom_bar()` layer on the same line after the `ggplot(metadata, ...)`
command from the previous exercise, and run the whole line.
What happens then? How do you interpret the result? Compare then to what
you get by running `table(metadata$age_bin)`.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

We get a full figure, yay!

```{r}
ggplot(metadata, aes(x = age_bin)) + geom_bar()
```

This bar plot indeed matches the counts in our data:

```{r}
table(metadata$age_bin)
```


:::

It might be hard to believe now but you now know almost everything about
making _ggplot2_ figures. 🙂 Everything else, all the amazing complexity
of [the possible _ggplot2_ figures](https://r-graph-gallery.com/ggplot2-package.html)
(really, look at that website!) comes down from combining these two
simple elements: aesthetics and geoms.

---

**Let's demonstrate the layering aspect a little bit more by adding more
layers to our figure! Use a `xlab()` function to add an x-axis label
element to your bar plot code above, again using the `+` operator.
Look at `?xlab` to see how to use this function in practice. What
changed about your figure now?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

Just as we did with _tidyverse_ `%>%` data transformation pipelines, as a
_ggplot2_ visualization pipeline gets more complex, it's a good idea to
introduce indentation so that each visualization steps is on its own line.

```{r}
ggplot(metadata, aes(x = age_bin)) +
  geom_bar() +
  xlab("Time period [years before present]")
```

By default, x and y axes are labeled by the data-frame column name
specified for that x or y axis in the `aes()` mapping. `xlab()` overrides
that default.

:::

---

**Now continue adding another layer with the `ylab()` function, giving
the plot proper y-axis label too.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ggplot(metadata, aes(x = age_bin)) +
  geom_bar() +
  xlab("Time period [years before present]") +
  ylab("Number of individuals")
```

:::

---

**Give your figure a proper main title using the function `ggtitle()`
(again, use help `?ggtitle` if you need to).**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ggplot(metadata, aes(x = age_bin)) +
  geom_bar() +
  xlab("Time period [years before present]") +
  ylab("Number of individuals") +
  ggtitle("Distribution of sample counts in each time period")
```

:::

---

**Although individual functions `xlab()`, `ylab()`, `ggtitle()` are useful,
oftentimes it's better to use a general function `labs()`. Look up its
documentation under `?lab`, then rewrite the code for your figure to use
only this function, replacing your uses of `xlab()`, `ylab()`, and `ggtitle()`
just with `labs()`. Note that the function has other useful arguments---go
ahead and use as many of them as you want.**

**Note:** Don't worry about making your figure cluttered with too much text,
this is just for
practice. In a real paper, you wouldn't use title or caption directly in a
_ggplot2_ figure, but it's definitely useful for work-in-progress reports at
meetings, etc.

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ggplot(metadata, aes(x = age_bin)) +
  geom_bar() +
  labs(
    y = "Number of individuals",
    x = "Time period [years before present]",
    title = "Distribution of sample counts in each time period",
    subtitle = "Counts for individuals in 1000 Genomes Project and MesoNeo data",
    caption = "Here is an optional caption for the figure, similarly to what
      you might do in a real scientific article as a more detailed description of
      of what is going on in the plot."
  )
```

:::

---

**You can see that the "present-day" bin completely overwhelms the number of
individuals in the oldest bins. This often happens with data which follow
a more or less exponential scale. A useful trick is adding a `+ scale_y_log()`
layer. You can probably guess what it does, so try adding it to your code!**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

Of course, this function transforms the y-axis on the logarithmic scale! Very
useful!

```{r}
ggplot(metadata, aes(x = age_bin)) +
  geom_bar() +
  labs(
    y = "Number of individuals",
    x = "Time period [years before present]",
    title = "Distribution of sample counts in each time period",
    subtitle = "Counts for individuals in 1000 Genomes Project and MesoNeo data",
    caption = "Here is an optional caption for the figure, similarly to what
      you might do in a real scientific article as a more detailed description of
      of what is going on in the plot."
  ) +
  scale_y_log10()
```

:::

---

**Does the order of adding layers with the `+` operator matter? Do a little
experiment to figure it out by switching the order of the plotting code
and then running each version in your R console!**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

No, generally, it doesn't. This is another very useful feature of _ggplot2_.
Remember that a huge important part of an R data science workflow is the interactive
communication with an R console. You can start with a little bit of code
which produces a basic outline of your figure and continue adding useful bits
and pieces to it, modifying the outcome as you go.

:::

---

It's useful to keep in mind that you can always pipe a data frame
into a `ggplot()` function using the `%>%` operator. After all, `%>%`
always places the "data frame from the left" as the first argument
in the `ggplot()` call, where the function expects it anyway.
I.e., instead of writing `ggplot(metadata, aes(...))`, you can also
do `metadata %>% ggplot(aes(...))`. This allows you to transform or summarize
data before plotting, which makes the combination of _tidyverse_ and _ggplot2_
infinitely stronger.

**As a refresher and for a bit more practice, `filter()` the metadata first
for individuals who are 10000 years or older, discarding the rest. Then pipe
this `filter()` result into your `ggplot()` code for the bar plot you
have created previously, keeping the plotting part exactly the same
(maybe skipping the log-scale).**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
metadata %>%
filter(age > 10000) %>%
ggplot(aes(x = age_bin)) +
  geom_bar() +
  labs(
    y = "Number of individuals",
    x = "Time period [years before present]",
    title = "Distribution of sample counts in each time period",
    subtitle = "Counts for individuals in 1000 Genomes Project and MesoNeo data",
    caption = "Here is an optional caption for the figure, similarly to what
      you might do in a real scientific article as a more detailed description of
      of what is going on in the plot."
  )
```

:::

Believe it or not, with this basic structure of _ggplot2_ visualization and
combining this with any other type of _tidyverse_ manipulation, filtering, and
(as we'll later see), summarization, you're ready to make any figure imaginable.
Let's take this one step further by taking a look at another type of visualization.

---

### Exercise 3: Distribution of a numerical variable

In the previous exercise you visualized a distribution of a categorical
variable, specifically the counts of samples in an age category.
Now let's look at continuous variables and this time consider not the
metadata table, but our `ibd_merged` table of IBD segments between pairs of individuals. Here it is again, as a refresher:

```{r}
head(ibd_merged)
```

The most interesting quantity we might want to analyse in the IBD context
is the `length` of IBD segments, which indicates the degree of genetic
relatedness between individuals. So let's take a look at the 
distribution of IBD lengths across our data set as an opportunity to introduce
new functionality and useful tricks of _ggplot2_.

#### `geom_histogram()` and `geom_density()`

You can visualize a distribution of values using the functions `geom_histogram()`
and `geom_density()`. Both are useful for similar things, each has its strengths
and weaknesses from a visualization perspective.

**Visualize the distribution of `length` values across all individuals in your
IBD data set, starting with these two general patterns, just like you did in
the previous `geom_bar()` example:**

1. `ggplot() + geom_histogram()`, and also
2. `ggplot() + geom_density()`.

**Of course, you have to specify `ibd_merged` as the first parameter of 
`ggplot()`, and fill in the `aes(x = length)` accordingly.
Try to use the knowledge you obtained in the
previous exercise! As a reminder, the `length` is expressed in units of
centimorgans, so immediately add an x-axis label clarifying these units to
anyone looking at your figure, either using `xlab()` or `labs(x = ...)`.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

Here's a histogram:

```{r}
ibd_merged %>%
  ggplot(aes(x = length)) +
  geom_histogram() +
  labs(x = "IBD length [centimorgans]")
```

Here's a density plot:

```{r}
ibd_merged %>%
  ggplot(aes(x = length)) +
  geom_density() +
  labs(x = "IBD length [centimorgans]")
```

:::

---

**To make our data exploration a little easier at the beginning, let's create
a new variable `ibd_eur` from the total `ibd_merged` table as its subset,
using the following _tidyverse_ transformations (both of them):**

1. `filter()` for `region_pair == "Europe:Europe"`
2. `filter()` for `time_pair == "present-day:present-day"`

**Save the result of both filtering steps as `ibd_eur`, and feel free to use
the `%>%` pipeline concept that you learned about previously.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

This should be easy for you now as budding _tidyverse_ data-munging experts!

```{r}
ibd_eur <-
  ibd_merged %>%
  filter(
    region_pair == "Europe:Europe" &
    time_pair == "present-day:present-day"
  )

ibd_eur
```

Note that I always like to indent individual steps of a data processing
pipeline so that each component is on its separate line. It helps with
readability a lot!

:::

---

**Now visualize the distribution of lengths in the reduced, Europe-only
IBD data set stored in the `ibd_eur` variable. Do this in two versions again,
using `geom_histogram()` or `geom_density()`, exactly like you did above.
Additionally, for each of the
two functions specify either `fill` or `color` (or both of them---experiment!) based on the
variable `country_pair`. Experiment with the looks of the result and find
your favourite (prettiest, cleanest, easiest to interpret, etc.).**

**Note:** Think about how powerful the _ggplot2_ framework is. You only need
to change one little thing (like swapping `geom_histogram()` for `geom_density()`) and keep everything else the same, and you get a completely
new figure! This is amazing for brainstorming figures for papers, talks, etc.

**How do you compare these different approaches based on the visuals
of the graphics they produce (readability, aesthetics, scientific usefulness,
etc.)? When does one seems more appropriate than other?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

1. Histogram version with `fill` set (I personally don't find the `color`
version of histograms useful at all so I'm not showing it here):

```{r}
ibd_eur %>%
  ggplot(aes(x = length, fill = country_pair)) +
  geom_histogram() +
  labs(x = "IBD segment length [centimorgans]")
```

2. Density versions:

- using `color` only:

```{r}
ibd_eur %>%
  ggplot(aes(x = length, color = country_pair)) +
  geom_density() +
  labs(x = "IBD segment length [centimorgans]")
```


:::

---

**When you create a density plot again, setting the `fill = country_pair`
aesthetic, you get overlapping densities which are hard to disentangle.
Does changing `geom_density()` for `geom_density(alpha = 0.25)` help a bit?
What does `alpha` do?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

`alpha` sets the transparency value to avoid geoms hiding behind each other,
but it doesn't help that much here. It's a useful parameter to keep in mind
though!

```{r}
ibd_eur %>%
  ggplot(aes(x = length, fill = country_pair, color = country_pair)) +
  geom_density(alpha = 0.5) +
  labs(x = "IBD segment length [centimorgans]")
```

:::

### Exercise 4: Categories in facets (panels)

Let's talk a little bit more about multiple dimensions of 
data (i.e., the number of columns in our data frame) and how we can
represent them. This topic boils down to the question of _"How many
dimensions of our data can we capture in a single figure and be it as
comprehensible as possible?"_ which is more and more important the more
complex (and interesting!) our questions become.

Looking at the figures of IBD length distributions again (both histograms
and densities), it's clear there's a lot of overlapping information that is
a little hard to interpret. Just look at it again:

```{r}
#| code-fold: true
ibd_eur %>%
  ggplot(aes(x = length, fill = country_pair)) +
  geom_histogram() +
  labs(x = "IBD segment length [centimorgans]")
```

In situations like this, _facetting_ can be very helpful.

**Faceting is easier to demonstrate than explain, so just go ahead, take
your IBD segment length histogram plot created by `geom_histogram()` above
and add yet another layer with `+ facet_wrap(~ country_pair)` at the end
of the plotting command. How do you interpret the result now? Is it easier
to read?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ibd_eur %>%
  ggplot(aes(x = length, fill = country_pair)) +
  geom_histogram() +
  labs(x = "IBD segment length [centimorgans]") +
  facet_wrap(~ country_pair)
```

:::

---

**What happens when you add these variations of the `facet_wrap()` layer instead?
When would you use one or the other?**

1. `+ facet_wrap(~ country_pair)`
2. `+ facet_wrap(~ country_pair, scales = "free_x")`
3. `+ facet_wrap(~ country_pair, scales = "free_y")`
4. `+ facet_wrap(~ country_pair, scales = "free")`


::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

I'll use this opportunity to show you a useful trick of storing a
_ggplot2_ result in a variable... because we can add to contents of
variables using the `+` operator thanks to the magic provided by _ggplot2_!

We can create our "base histogram" like this, storing it in the variable
`p_histogram`---**try it yourself, please!**

```{r}
p_histogram <- ggplot(ibd_eur, aes(length, fill = country_pair)) +
  geom_histogram() +
  labs(x = "IBD segment length [centimorgans]")
```

##### This is our base histogram (very hard to read)

We you type out just the variable, you get the figure!

```{r}
p_histogram
```

##### 1. option

```{r}
p_histogram + facet_wrap(~ country_pair)
```

And here is what you get with various `facet_wrap()` options:

##### 2. option -- each facet has its own x-axis scale

```{r}
p_histogram + facet_wrap(~ country_pair, scales = "free_x")
```

##### 3. option -- each facet has its own y-axis scale

```{r}
p_histogram + facet_wrap(~ country_pair, scales = "free_y")
```

##### 4. option -- each facets has its own x- and y-axis scales

```{r}
p_histogram + facet_wrap(~ country_pair, scales = "free")
```

All approaches have their benefits. In general, personally, I think that
figures which show the same statistic across different categories should
generally use the same scales on both x- and y-axes, because this makes
each facet comparable to each other.

But there are pros and cons to doing all of these in different situations!

:::


---


### Exercise 5: Limiting axis scales

**Sometimes when visualizing distributions of numerical variables (histograms,
densities, any other geom we'll see later), the distribution is too widely
spread out to be useful. We can mitigate this using these two additional
new layers which restrict the range of the x-axis of our figures:**

1. `+ xlim(1, 7)`
2. `+ coord_cartesian(xlim = c(1, 7))`

**Take your density distribution ggplot of lengths of IBD segments again
(you can use my code right below) and add 1. or 2. option above. What's the difference between both possibilities?**

```{r}
#| eval: false
ibd_eur %>%
    ggplot(aes(x = length, color = country_pair)) +
    geom_density() +
    labs(x = "IBD segment length [centimorgans]")
```


::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

Here are the two versions of restricting the x-axis scale. Note that they
differ only by the command used at the 5th line!

1. `coord_cartesian()` performs so-called "soft clipping" -- it restricts the
"viewpoint" on the entire data set:

```{r}
#| code-line-numbers: true
ibd_eur %>%
  ggplot(aes(x = length, color = country_pair)) +
  geom_density() +
  labs(x = "IBD segment length [centimorgans]") +
  coord_cartesian(xlim = c(1, 7))
```

2. `xlim()` does "hard clipping" -- it effectively removes the points outside
of the given range. This usually isn't what we want, so I personally rarely
if ever use this:

```{r}
#| code-line-numbers: true
ibd_eur %>%
  ggplot(aes(x = length, color = country_pair)) +
  geom_density() +
  labs(x = "IBD segment length [centimorgans]") +
  xlim(1, 7)
```

Admittedly they both give the same result in this situation, but that's not always the case, so be careful! I personally usually use the option 1., just
to be safe.

:::

---

**Is there evidence of some countries sharing more IBD sequence
_within themselves_. If that's the case, it could be a result of a stronger
population bottleneck. The previous density plot is a little too busy to
be able to see this, so filter your `ibd_eur` table for rows in which
`country1` is the same as `country2`, and then %>% pipe it into the
usual _ggplot2_ density command.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ibd_eur %>%
  filter(country1 == country2) %>%
  ggplot(aes(length, color = country_pair)) +
  geom_density() +
  labs(x = "IBD segment length [centimorgans]") +
  xlim(0, 10)
```

It looks like there are more IBD sequence shared within Finland, which is
consistent with a
[previously established hypothesis](https://en.wikipedia.org/wiki/Finnish_heritage_disease)
that ancestors of present-day Finns experienced a strong popularion bottleneck.

:::

---

#### `geom_boxplot()` and `geom_violin()`

**Let's introduce another useful geom for visualization of this type of
numerical data, the boxplot! You can visualize boxplots with the function
`geom_boxplot()`, and calling (in our case right here)
`aes(x = country_pair, y = length, color = country)` within the `ggplot()`
function. Adjust your previous density example according to these instructions
to show boxplots instead!**

**When you have your boxplots, swap out `geom_boxplot()` for `geom_violin()`.**

**Note:** Again, notice how super modular and customizable
_ggplot2_ for making many different types of visualizations. We just had
to swap one function for another.

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ibd_eur %>%
  filter(country1 == country2) %>%
  ggplot(aes(x = country_pair, y = length, color = country_pair)) +
  geom_boxplot() +
  labs(y = "IBD segment length [centimorgans]") +
  ylim(0, 10)
```

```{r}
ibd_eur %>%
  filter(country1 == country2) %>%
  ggplot(aes(x = country_pair, y = length, color = country_pair)) +
  geom_violin() +
  labs(y = "IBD segment length [centimorgans]") +
  ylim(0, 10)
```

:::


I'm personally a much bigger fan of the violin plot for this kind of purpose,
because I feel like boxplot hides too much about the distribution of the data.

**On that note, another useful trick we haven't mentioned yet is that we can
plot multiple geoms from the same data! You're already quite familiar with
the _ggplot2_ concept of layering functions on top of each other using the `+`
operator. What happens when you add `+ geom_jitter(size = 0.3, color = "darkgray")` (again at the end of the code which produces violin)?
How do you read the resulting figure?**

**Note:** We specified `color = "darkgray"`, which effectively overrides
the `color = country_pair` assigned "globally" in the `ggplot()` call
(and which normally sets the color for every single geom we would add to our figure).

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

`geom_jitter()` is extremely useful to really clearly see the true distribution
of our data points (which is often otherwise hidden by the "summarizing geoms",
such as boxplot above -- one reason why I'm not a fan of boxplots, as I 
mentioned!):

```{r}
ibd_eur %>%
  filter(country1 == country2) %>%
  ggplot(aes(x = country_pair, y = length, color = country_pair)) +
  geom_violin() +
  geom_jitter(size = 0.3, color = "darkgray") +
  labs(y = "IBD segment length [centimorgans]") +
  ylim(0, 10)
```

:::

---

**What do you think happens when you swap the order of `geom_violin()`
and `geom_jitter(...)` in your previous solution? Why does the order matter?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

_ggplot2_ arranges elements on a canvas literally like a painter would,
if an element is added and another element is added later, the latter can
(partially) overlap the former:

```{r}
ibd_eur %>%
  filter(country1 == country2) %>%
  ggplot(aes(x = country_pair, y = length, color = country_pair)) +
  geom_jitter(size = 0.3, color = "darkgray") + # we flipped the order
  geom_violin() +                               # of these two geoms
  labs(y = "IBD segment length [centimorgans]") +
  ylim(0, 10)
```

Of course, this can also be intentional sometimes! It all depends on the
aesthetic you're going for as a visualization artist. :)

:::



### Intermezzo: Saving _ggplot2_ figures to PDF

Looking at figures in RStudio is useful for exploratory analysis, but when
we want to use a figure in a paper or a presentation, we need to save it to
disk. Doing this is very easy using the function `ggsave()`.

**There are generally three options:**

1. **Simply running a _ggplot2_ command, and calling `ggsave()` right after
to save a single figure.** Something like this:

```{r}
#| eval: false
# we first plot a figure by running this in the R console:
ggplot(ibd_eur, aes(length, fill = country_pair)) +
  geom_histogram() +
  labs(x = "IBD segment length [centimorgans]")

# this then saves the last figure to a given file
ggsave("~/Desktop/ibd_length_histogram.pdf")
```

2. Or, sometimes in a larger script (like we'll see later), it's useful
to actually **save a ggplot to a variable (yes, you can do that!), and then
give that variable to `ggsave()` as its second parameter**.
This is useful when you want to first
create multiple plots, store them in their own variables, then save them
to files at once. Something like this:

```{r}
#| eval: false
# we first create a histogram plot
p_histogram <- ggplot(ibd_eur, aes(length, fill = country_pair)) +
  geom_histogram() +
  labs(x = "IBD segment length [centimorgans]")

# then we create a density plot
p_density <- ggplot(ibd_eur, aes(length, fill = country_pair)) +
  geom_density(alpha = 0.7) +
  labs(x = "IBD segment length [centimorgans]")

# note that if you type the variable into your R console (try this!),
# you will get your figure visualized:
p_histogram

# we can save figures stored in variables like this
ggsave("~/Desktop/ibd_length_histogram.pdf", p_histogram)
ggsave("~/Desktop/ibd_length_density.pdf", p_density)
```

3. **Save multiple figures into a single PDF.** This requires a different
approach, in which we first initialize a PDF file, then we have to print(!)
individual variables with each _ggplot2_ figure, and then close that PDF
file. Yes, this is confusing, but don't worry about the technical details of
it! Just take this as a recipe you can revisit later if you need to.

```{r}
#| eval: false
# this creates an empty PDF
pdf("~/Desktop/multiple_plots.pdf")

# we have to print(!) individual variables with each plot
print(p_histogram)
print(p_density)

# this closes the PDF
dev.off()
```

**I don't think it's very interesting to create a specific exercise for it.
Just try running these three examples on your own, and see what happens!
We'll be using these techniques a lot in the next session, so just keep
them in mind for now.**

**Note:** Take a peek at `?ggsave` and `?pdf` help pages to see
customization options, particularly on the sizes of the PDFs.
Also, how would you create a PNG images with the figures? Use the help pages!


### Exercise 6: Relationship between two variables

#### `geom_point()`

We'll now look at creating another type of visualization which is extremely
useful and very common: a [scatter plot](https://en.wikipedia.org/wiki/Scatter_plot)!

But first, let's do some more data summarization so that we can demonstrate
building of scatter plots using _ggplot2_. I actually created a new data set
for you, using the original IBD segments table.

It contains, for a given pair of individuals in the data set, several summary
statistics about their level of IBD sharing across the entire genome
(not just chromosome 21). Specifically, you will now focus on three columns
of the new table:

- `n_ibd` --- the total number of IBD segments between those two individuals
- `total_ibd` --- the total amount of genome in IBD between those two individuals
- `rel` --- indicator of the relatedness between those two individuals

**Read this table using this command:**

<!--  At this point, this should all be -->
<!-- a piece of cake for you! -->

<!-- The previous figures showed distributions of individual segment lengths -->
<!-- on present-day individuals only. Another useful quantity is the number of -->
<!-- segments shared between a pair of individuals and the total number of IBD -->
<!-- sequence between those pairs. -->

<!-- **Practice your knowledge of `group_by()` and `summarize()` and apply them -->
<!-- to our huge `ibd_segments` table, to achieve the following:** -->

<!-- 0. Start with our original big IBD table `ibd_segments` (not the smaller `ibd_eur`!) -->

<!-- 1. `filter()` for rows in which `region_pair == "Europe:Europe"` (only pairs -->
<!-- within Europe, just as we did before), -->

<!-- 2. `filter()` for rows in which `length > 10` (only long IBD segments) -->

<!-- 3. `filter()` for rows in which `age_bin1 == age_bin2` (only individuals from -->
<!-- the same age bin), -->

<!-- 4. `group_by()` to create groupings based on `sample1`, `sample2`, -->
<!-- `country_pair`, `region_pair`, and `time_pair`, -->

<!-- 5. `summarize()` to create two new quantities `n_ibd = n()` and `total_ibd = sum(length)` (total number of long IBD segments and their total sequence for each -->
<!-- of individuals). -->

<!-- **Save the result of your `%>%` pipeline implementing the steps 1.-3. above in -->
<!-- a variable `ibd_sum`.** -->


<!-- ::: {.callout-note collapse="true" icon=false} -->
<!-- #### Click to see the solution -->

<!-- Again, once you know the functions `filter()`, `group_by()`, and `summarize()`, -->
<!-- implementing the pipeline is just a matter of sequencing them together as if -->
<!-- you were building a LEGO set! -->

<!-- ```{r} -->
<!-- ibd_sum <- -->
<!--   ibd_merged %>% -->
<!--   filter( -->
<!--     length > 10 &          # only long IBD segments -->
<!--     age_bin1 == age_bin2   # only IBD pairs within the same time window -->
<!--   ) %>%  -->
<!--   group_by(sample1, sample2, rel, country_pair, region_pair, time_pair) %>% -->
<!--   summarize(n_ibd = n(), total_ibd = sum(length)) -->

<!-- ibd_sum -->
<!-- ``` -->

<!-- I hope you're all appreciating the awesome data science powers you've already -->
<!-- learned to do this! This would require a crazy amount of work with standard -->
<!-- base R and might be very hard (or downright impossible) to do with Excel. -->

<!-- Believe me that if you understand this particular chunk of code (even if -->
<!-- you needed help putting it together) you know the majority of _tidyverse_ -->
<!-- work needed to answer complex data science questions. -->

<!-- ::: -->

<!-- --- -->

<!-- **Your work so far has been based on a small subset of the overall IBD -->
<!-- data, just for chromosome `r unique(ibd_segments$chrom)`. To move to more complex -->
<!-- and interesting visualization, execute this command which reads the `ibd_long` -->
<!-- just like you created it yourself, except based on all chromosomes in the -->
<!-- entire genome. We'll be working on that from now on.** -->

```{r}
ibd_sum <- read_tsv("http://tinyurl.com/simgen-ibd-sum")
```


---

With our summarized table `ibd_sum` ready, we're ready to create our
scatter plot. When we worked with `geom_histogram()`, `geom_density()`, and
`geom_bar()`, we only needed to do something like this:

```{r}
#| eval: false
df %>%
  ggplot(aes(x = COLUMN_NAME)) +
  geom_FUNCTION()
```

Because scatter plot is two-dimensional, we have to specify
not only the `x` but also the `y` in the `aes()` aesthetic mapping... and with
that bit of information, I think you already know how to do this based
on your knowledge of basic _ggplot2_ techniques introduced above!

**Use the function `geom_point()` to visualize the scatter plot of
`x = total_ibd` against `y = n_ibd`. We know that related individuals share
a large amounts of IBD sequence between each other. As a reminder, the entire
human genome spans about 3000 centimorgans. Can you guess from your figure
which pairs of individuals (a point on your figure representing one such pair)
could be potentially closely related?**

**Note:** Don't worry about this too much, just take a guess. Below we'll look
at this question more closely.

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

It looks like some pairs of individuals share almost their entire genome in
IBD, around the whole 3 gigabases of sequence!

We'll work on clarifying that below.

```{r}
ibd_sum %>%
  ggplot(aes(x = total_ibd, y = n_ibd)) +
  geom_point()
```

:::

---

**From the introduction of histograms and densities, you're already familiar
with modifying the `aes()` aesthetic of a geom layer using the `color`
argument of `aes()`. Right now your dots are all black, which isn't super
informative. Luckily, your IBD table has a "mysterious" column called
`rel`. What happens when you `color` points based on the values in this
column inside the `aes()` function (i.e., set `color = rel`)? Similarly,
what happens when you set `shape = rel`?**

**Look at a figure from [this huge](https://www.nature.com/articles/s41588-023-01582-w/figures/3)
study of IBD patterns and their relationship to the degree of relatedness
between individuals? This is completely independent data set and independent
analysis, but do you see similarities between this paper and your figure?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ibd_sum %>%
  ggplot(aes(x = total_ibd, y = n_ibd, color = rel)) +
  geom_point()
```

:::

I have to say I kind of hate that "none" (basically a baseline background
signal)
is given a striking purple colour (and I love purple). I wouldn't want that
in my paper, and we will show a potential solution to this soon!
For now, let's keep going.

---


**To get a bit closer to a presentation-ready figure, use the
`+ labs()` layer function to properly
annotate your `x` and `y` axes, and give your figure a nice `title` too.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ibd_sum %>%
  ggplot(aes(x = total_ibd, y = n_ibd, color = rel)) +
  geom_point() +
  labs(x = "total IBD sequence", y = "number of IBD segments",
       title = "Total IBD sequence vs number of IBD segments",
       subtitle = "Both quantities computed only on IBD segments 10 cM or longer")
```

:::

### Exercise 6: Multiple data sets in one plot

So far we've always followed this pattern of _ggplot2_ usage, to borrow our
very first `geom_histogram()` example just for the sake of explanation:

```{r}
ggplot(ibd_segments, aes(length)) + geom_histogram()
```

**However, you can also do this and it will give the same result (try it--
it will come in handy many times in the future!)**:

```{r}
ggplot() +
  geom_histogram(data = ibd_segments, aes(length))
```

In other words, each `geom_<...>()` function accepts both `data` data frame
and also its own `aes()` aesthetic and mapping function. You could think of
this as each individual geom having a "self-contained" `ggplot()` function's
ability.

**This can be very useful whenever you want to plot different geoms
with their own aesthetic parameters (colors, shapes, sizes)!** Let's
work through an exercise to explain this better, step by step, to make our
IBD scatter plot figure much more nicer to look at.

---

Above I complained a little bit how the "none" related data points are
given just as strikingly colorful points as data points of interest. What we
want to establish is different visual treatments of different kinds of data.
Here's one solution to do this (which will also demonstrate a very very
useful _ggplot2_ trick).

**First, take your `ibd_sum` table and create two versions of it
using `filter()` like this:**

1. `ibd_unrel <- filter(ibd_sum, rel == "none")`
2. `ibd_rel <- filter(ibd_sum, rel != "none")`

The first table contains only those pairs of individuals with missing information
about relatedness, the second one contains only those rows with this information
present.

```{r}
#| echo: false
ibd_unrel <- filter(ibd_sum, rel == "none")
ibd_rel <- filter(ibd_sum, rel != "none")
```

---

**Now visualize the `ibd_unrel` (i.e., IBD summaries between unrelated individuals)
table with the following code:**

```
ggplot() +
  geom_point(data = ibd_unrel, aes(x = n_ibd, y = total_ibd), color = "lightgray")
```

**Note:** The `ggplot()` in this case has no arguments, just like we
showed in the two versions of `geom_histogram()` above. Everything is in
the `geom_point()` call!

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ggplot() +
  geom_point(data = ibd_unrel, aes(x = n_ibd, y = total_ibd), color = "lightgray")
```

:::

---

**Now visualize the scatter plot of pairs of individuals which are known to be
related, `ibd_rel`:**

```
ggplot() +
  geom_point(data = ibd_rel, aes(x = n_ibd, y = total_ibd, color = rel))
```

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
ggplot() +
  geom_point(data = ibd_rel, aes(x = n_ibd, y = total_ibd, color = rel))
```

:::

---

**Compare the two figures you just created, based on `ibd_rel` and `ibd_unrel`
to the complete figure based on `ibd_sum` you created above. You can see
that these to plots together are showing the same data.**

---

You have now plotted the two scatter plots separately, but here comes a very
cool aspect of _ggplot2_ I wanted to show you. **You can combine the two figures,
each plotting a different data frame, into one figure by writing _ggplot2_
code which has both `geom_point()` calls for `ibd_unrel` and `ibd_rel` in
the same command!**

**Do this now and to make the
figure even nicer (and even publication ready!), add proper `x` and `y` labels,
overall plot `title` as well as `subtitle` (all using the `labs()` layer
functions) as well as adjust the legend title (just like you did with the
`guides()` layer function above).**

**In the `geom_point()` command plotting unrelated individuals (`ibd_unrel`),
in addition to setting `color = "lightgray"`, set also `size = 0.75`.**

**Hint:** If it still isn't clear, don't worry. Here's a bit of help of what
your plotting code should look like:

```
ggplot() +
  geom_point(<YOUR CODE PLOTTING UNRELATED IBD PAIRS>)
  geom_point(<YOUR CODE PLOTTING RELATED IBD PAIRS>) +
  labs(<x, y, title, and subtitle text>) +
  guides(color = guide_legend("relatedness"))
```


::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

Here's the complete solution to this exercise:


```{r}
ibd_unrel <- filter(ibd_sum, rel == "none")
ibd_rel <- filter(ibd_sum, rel != "none")

ggplot() +
  geom_point(data = ibd_unrel, aes(x = total_ibd, y = n_ibd), color = "lightgray", size = 0.75) +
  geom_point(data = ibd_rel, aes(x = total_ibd, y = n_ibd, color = rel)) +
  labs(
    x = "total IBD sequence",
    y = "number of IBD segments",
    title = "Total IBD sequence vs number of IBD segments",
    subtitle = "Both quantities computed only on IBD segments 10 cM or longer"
  ) +
  guides(color = guide_legend(title = "relatedness"))
```

:::

---


**As a bonus, and to learn another parameter of `aes()` useful for scatter plots,
in addition to setting `color = rel` (which sets a color of each point based
on the relatedness value), add also `shape = rel`. Observe what happens
when you do this!**

**Finally, I'm not personally fan of the default grey-filled background of ggplots.
I like to add `+ theme_minimal()` at the end of my plotting code of many of
my figures, particularly when I'm creating visualizations for my papers.
Try it too!**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

Here's the complete solution to this exercise:


```{r}
ibd_unrel <- filter(ibd_sum, rel == "none")
ibd_rel <- filter(ibd_sum, rel != "none")

ggplot() +
  geom_point(data = ibd_unrel, aes(x = total_ibd, y = n_ibd), color = "lightgray", size = 0.75) +
  geom_point(data = ibd_rel, aes(x = total_ibd, y = n_ibd, color = rel, shape = rel)) +
  labs(
    x = "total IBD sequence",
    y = "number of IBD segments",
    title = "Total IBD sequence vs number of IBD segments",
    subtitle = "Both quantities computed only on IBD segments 10 cM or longer"
  ) +
  guides(
    color = guide_legend(title = "relatedness"),
    shape = guide_legend(title = "relatedness")
  ) +
  theme_minimal()
```

:::

---




## Bonus Part 2: Practicing visualization on time-series data

### Exercise 8: Getting fresh new new data

We've spent a lot of time looking at metadata and IBD data, demonstrating
various useful bits of techniques from across _tidyverse_ and _ggplot2_.
There are many more useful things to learn, but let's switch gears a little
bit and look at a slightly different kind of data set, just to keep things
interesting.

**First, create a new script and save it as `f4ratio.R`.**

**Then add the following couple of lines of setup code, which again loads
the necessary libraries and also reads a new example data set:**

```{r}
library(dplyr)
library(readr)
library(ggplot2)

f4ratio <- read_tsv("https://tinyurl.com/simgen-f4ratio", show_col_types = FALSE)
```

This data set contains results from a huge population genetics simulation
study I recently did, and contains estimates of the proportion of Neanderthal
ancestry in a set of European individuals spanning the past 50 thousand years.

Here's a brief description of the contents of the columns:

1 `individual`, `time` -- name and time (in years before present) of an individual
2. `statistic` -- a variant of an estimator of Neanderthal ancestry
3. `proportion` -- the proportion of Neanderthal ancestry estimated by `statistic`
4. `rate_eur2afr` -- the amount of ancient migration (gene flow) from
Europe to Africa in a given scenario
5. `replicate` -- the replicate number of a simulation run (each scenario was
run multiple times to capture the effect of stochasticity)

```{r}
head(f4ratio)
```

In other words, each row of this table contains the `proportion` of Neanderthal
ancestry estimated for a given `individual` at a given `time` in some scenario
specified by a given set of other parameters (`rate_afr2afr`, `rate_eur2afr`,
`replicate`).

**In this exercise, I will be asking you some questions, the answers to
which you should be able to arrive to by the set of techniques from
_tidyverse_ you've already learned, and visualize them using the _ggplot2_
functions you already know. When a new function will be needed, I will be
dropping hints for you along the way!**

**This exercise is primarily intended for you to practice more _ggplot2_
visualization, because I think this is the most crucial skill in doing
any kind of data analysis. Why? The faster you can get from an idea to
a visualization of that idea, the faster you'll be able to do research.**

Yes, we could do this on the MesoNeo metadata or the IBD data too, but
I think it would be too boring to stay with the same set of tables for this
long. :)

Let's go!

---

**Use `geom_point()` to plot the relationship between `time` (`x` axis`)
and `propotion` of Neanderthal ancestry (`y` axis). Don't worry about any
other variables just yet (colors, facet, etc.).**

**Note:** The `f4ratio` table is too huge to plot every single point.
Before you `%>%` pipe it into `ggplot()`, _subsample_ the rows to just
10% of it subset using this trick:

```
f4ratio %>%
  sample_frac(0.1) %>% # subsample 10% fraction of the data
  ggplot(aes(...)) +
    geom_point()
```

**Hint:** Look at `?sample_frac` but also `?sample_n` and check out their
parameters. They are incredible useful for any kind of data exploration!

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
f4ratio %>%
sample_frac(0.1) %>%
ggplot(aes(time, proportion)) +
  geom_point()
```

:::

---

**Your plot now mashes together `proportion` values from two different
`statistic` estimators. Fix this by facetting the plot, which will visualize
both statistics in two separate panels. For the time being, keep doing the
10% subsampling step.**

**Hint:** You can do this by adding `+ facet_wrap(~ statistic)` to your previous code.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
f4ratio %>%
sample_frac(0.1) %>%
ggplot(aes(time, proportion)) +
  geom_point() +
  facet_wrap(~ statistic)
```

:::

---

**Take a look at the values on the x-axis of your figures. Purely for
aesthetics, it is customary to visualize time going from oldest (left) to
present-day (right). But currently the time direction is reversed.
Fix this by adding a new layer with `+ scale_x_reverse()`!**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
f4ratio %>%
sample_frac(0.1) %>%
ggplot(aes(time, proportion)) +
  geom_point() +
  facet_wrap(~ statistic) +
  scale_x_reverse()
```

:::

---

**Whenever we visualize some biological statistics, we often want to compare
their values to some kind of baseline (or a null hypothesis) value. In our
case we have two:**

1. baseline of "no Neanderthal ancestry" at y-axis value 0,
2. baseline of "known proportion of Neanderthal ancestry" today, which is
about 3% (so y-axis value of 0.03).

**Add those two baselines to your figure using a new function
`geom_hline()` (standing for "horizontal line"). Like this:**

```
<YOUR TIDYVERSE AND GGPLOT2 CODE ABOVE> +
  geom_hline(yintercept = 0.03, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "black", linetype = "dotted")
```

**Note:** You can immediately see the possibilities for adjusting the visuals
like colors or linetype. Feel free to experiment with this!

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
f4ratio %>%
sample_frac(0.1) %>%
ggplot(aes(time, proportion)) +
  geom_point() +
  facet_wrap(~ statistic) +
  scale_x_reverse() +
  geom_hline(yintercept = 0.03, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "blue", linetype = "dotted")
```

:::

---

There's clearly a huge amount of statistical noise, represented particularly
in the "indirect f4-ratio" figure. And we're not even plotting all
points because there's just too many of them!

**Additionally, what I cared about in this particular case (during my PhD)
were not values for each individual person, but about the _statistical trend_
in my data. An extremely important function for this is `geom_smoot()`,
which fits a line through the data, capturing this trend.**

**In this next part, remove the subsampling done by `sample_frac()`, and
also replace your `geom_point()` call in your figure with
this: `+ geom_line(stat = "smooth", se = FALSE)`. What happens when you do
this? Read `?geom_line` to figure out what `se = FALSE` means. Why do we want
to do this?**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
f4ratio %>%
ggplot(aes(time, proportion)) +
  geom_line(stat = "smooth", se = FALSE) +
  facet_wrap(~ statistic) +
  scale_x_reverse() +
  geom_hline(yintercept = 0.03, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "blue", linetype = "dotted")
```

:::

---

**Well, we removed a lot of noise but we also removed most of the interesting
signal! This is because we are now basically "smoothing" (averaging) over
a very important variable: the amount of migration from Europe to Africa
(which was the
[primary research interest for me at the time](https://www.pnas.org/doi/10.1073/pnas.1814338116)). Let's take a look
at our table again:**

```{r}
head(f4ratio)
```

**What we want is to plot not just a single line across every single scenario,
but we want to plot one line for each value of our variable of interest,
`rate_eur2afr`.**

**Let's add the variable into play by adding `color = rate_eur2afr` into your
`aes()` call in the `ggplot()` function (i.e., in addition to the `x = time`
and `y = proportion` you already got there).**

**Note:** You'll get a warning which we will fix in the next step, don't worry!

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
f4ratio %>%
ggplot(aes(time, proportion, color = rate_eur2afr)) +
  geom_line(stat = "smooth", se = FALSE) +
  facet_wrap(~ statistic) +
  scale_x_reverse() +
  geom_hline(yintercept = 0.03, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "blue", linetype = "dotted")
```

Don't worry about the warning, we'll fix this soon!

:::

---

The reason we got a warning here is a slightly confusing behavior of the
smoothing function `geom_smooth()`. Because `rate_eur2afr` is a numerical
variable, the smoothing function cannot plot a single smoothed line for
each of its values, because there could be infinite numnber of them.

**Fix the problem by adding `group = rate_eur2afr` right after your
`color = rate_eur2afr` you added in the previous exercise. This explicitly
instructs `geom_smooth()` to treat individual values of `rate_eur2afr` as
discrete categories.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
f4ratio %>%
ggplot(aes(time, proportion, color = rate_eur2afr, group = rate_eur2afr)) +
  geom_line(stat = "smooth", se = FALSE) +
  facet_wrap(~ statistic) +
  scale_x_reverse() +
  geom_hline(yintercept = 0.03, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "blue", linetype = "dotted")
```

:::

---

**Let's finalize our figure [for publication](https://www.pnas.org/doi/10.1073/pnas.1814338116#fig02)!
Use `labs()` to add proper `x`, `y`, and `title` descriptions (something like
"Neanderthal ancestry proportion", "years before present", and
"Expected Neanderthal ancestry trajectories").**

**Don't forget about the title of the legend too, which you can adjust by 
`+ guides(color = guide_legend("EUR -> AFR\nmigration"))`.**

**Finally, add my favourite `+ theme_minimal()` at the end.**

::: {.callout-note collapse="true" icon=false}
#### Click to see the solution

```{r}
f4ratio %>%
ggplot(aes(time, proportion, color = rate_eur2afr, group = rate_eur2afr)) +
  geom_line(stat = "smooth", se = FALSE) +
  facet_wrap(~ statistic) +
  scale_x_reverse() +
  geom_hline(yintercept = 0.03, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "blue", linetype = "dotted") +
  labs(x = "Neanderthal ancestry proportion",
       y = "years before present",
       title = "Expected Neanderthal ancestry trajectories",
       subtitle = "Results from a set of population genetic simulations") +
  guides(color = guide_legend("EUR -> AFR\nmigration")) +
  theme_minimal()
```

:::

**I hope that you now appreciate how adding little bits and pieces, one
step at a time (and each step being very tiny and easily understandable!),
makes it possible to create beautiful informative visualizations of any
kind of tabular data!**

### Exercise 9: Combining figures

So, here's an extremely cool revelation. Do you remember how we went through
various data types and object classes in the R bootcamp? Numeric, logical,
character types, etc?

_ggplot() functions actually do return a (hidden) value, which is a normal
object like any other R value!_ The fact that we immediately see a plot
visualized is "just" a side effect that R provides for convenience! After
all, when we write plotting code, we usually want to see the plot, right?

What does this mean? Well, first of all, consider this bit of code:

```{r}
p1 <- ggplot()
```

Weird, huh? Why would you do that? **When you type `p1` in your R console
you get the usual (empty) figure!**

```{r}
p1
```

Of course, we can also create (and save in a variable) a full figure like this:

```{r}
p1 <- ggplot(ibd_sum, aes(total_ibd)) + geom_density() + ggtitle("Normal x-axis scale IBD total sum")
p1
```

We could also create and save another figure, this time a density on a log
scale, just by adding a new layer to the `p1` variable plot:

```{r}
p2 <- p1 + scale_x_log10() + ggtitle("Logarithmic x-axis scale IBD total sum")
p2
```

This is extremely useful for making composite figures built out of individual
panels (like you often see in scientific figures as panels **A**, **B**, **C**,
etc.). The way we can do this is using a helper R package _cowplot_ (don't
ask me about its name, I have no idea! :)) and its function `plot_grid()`. **Check
this out:**

```{r}
library(cowplot)

plot_grid(p1, p2, labels = c("A", "B"))
```

This is an extremely useful pattern because it basically eliminates the need
for manually pasting figures in a graphics editor (which defies the purpose
of reproducibility!). The sky is actually a limit, because you can place
other `plot_grid()` calls within another `plot_grid()`, composing figures
of various levels of complexity. Check out
[a figure from one of my papers](https://github.com/bodkan/archaic-ychr/blob/master/figures/fig1.pdf),
each panel is an independently made _ggplot2_ figure, all pasted together
with `plot_grid()`. 

For instance, if we make `p3` like this:

```{r}
# create a silly scatter plot of geographical coordinates of our samples
# (do you see an outline of Europe? :))
p3 <-
  metadata %>%
  filter(continent == "Europe") %>%
  ggplot(aes(longitude, latitude)) +
    geom_point() +
  ggtitle("A silly toy map of Europe drawn using locations of metadata points")
p3
```

**You could combine them like this** (this is just an example, don't read too
much into the meaning of putting these particular plots together).

```{r}
plot_grid(
  p3,
  plot_grid(p1, p2, labels = c("B", "C")),
  nrow = 2,
  labels = "A"
)
```

## Advanced exercises

If you've already been introduced to _ggplot2_ before and the exercises
above were too easy for you, here are a couple of additional challenges:

1. **Take a look at how the
[_ggridges_ package](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html) can potentially improve the readability of
many density distributions when plotted for several factors ordered along
the y-axis.** For instance, how could you use this package to visualize
the (admittedly practically useless, but still) distributions of longitudes
(on the x-axis) for all samples in our `metadata`, ordered along the y-axis
by the country? Note that you might also need to learn about ordering
factors using the R package [_forcats_](https://forcats.tidyverse.org/articles/forcats.html).

As an extra challenge, try to make this figure as beautiful, colorful,
and magazine publication-ready using _ggplot2_ theming

2. **Study how you could turn our visualization of total IBD vs number of
IBD segments as a function of the degree of relatedness into a proper
interactive figure using [_ggplotly_](https://plotly.com/ggplot2/).**
For instance, make an interactive figure which will show the exact
count and total IBD, as well as the names of individuals upon hovering
over each data point with a mouse cursor.

## "Take home exercise"

**Take your own data and play around with it using the concepts you
learned above.** If the data isn't in a form that's readily readable
as a table with something like `read_tsv()`, please talk to me! _tidyverse_
is huge and there are packages for munging all kinds of data. I'm
happy to help you out.

**Don't worry about _"getting somewhere"_. Playing and experimenting
(and doing silly things) is the best way to learn.**

**For more inspiration on other things you could do with your data,
take a look at the [_dplyr_ cheatsheet](https://bodkan.net/simgen/cheatsheets/dplyr.pdf)
and the [_ggplot2_ cheatsheet](https://bodkan.net/simgen/cheatsheets/dplyr.pdf). You can also get an inspiration in the huge
[gallery of _ggplot2_ figures online](https://r-graph-gallery.com/)!**


## Additional resources

As with our introduction to _tidyverse_, there's so much more to _ggplot2_
than we had a chance to go through. I hope we'll get through some spatial
visualization features of R (which are tightly linked with _tidyverse_ and
_ggplot2_ in the modern R data science landscape), but there's much more
I wish we had time to go through.

Here are a couple of resources for you to study after the conclusion of
the workshop, or even during the workshop itself if you find yourself with
more time and more energy to study.

1. Factors are both a curse and blessing. When dealing with visualization,
often a curse, because our factors are rarely ordered in the way we want.
If our factors are numerical but not straightforward numbers (like our `age_bin`
column above), their order in _ggplot2_ figures is often wrong. _forcats_
is an incredible package which helps with this. Read about it in
[this vignette](https://forcats.tidyverse.org/articles/forcats.html) and then
experiment with using it for figures in this session which featured factors.

2. _cowplot_ is another R package that's very useful for complex composite
figures. We touched upon it in this session, but only extremely vaguely.
Study [this introduction](https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html)
to learn much more about its features! You will never need Inkscape again.

3. _ggridges_ is a very cool package to visualize densities across multiple
categories. It can often lead to much more informative figures compared to
overlapping `geom_density()` plots we did in our session. Here's a great
[official introduction](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html)
to the package.

