---
title: "Neanderthal ancestry throuhg time"
author: "WRITE YOUR NAME HERE :)"
date: now
format: html
self-contained: true
---

# Basic processing

## Load required packages

```{r}
# add required code here
```

## Read in data from the internet

```{r}
# add required code here
```

## Inspect the format of the table

First couple of rows of the data:

```{r}
# add required code here
```

A `glimpse` at the column values:

```{r}
# add required code here
```

# Visualizing data

## Visualize the 10% of the data points

```{r}
# add required code here (it used the `sample_frac()` function)
```

## Fit a smoothed out line through all parameter combinations

```{r}
# add required code here which plotted the beautiful finalized figure
# of smoothed lines of Neanderthal ancestry trajectories
```

# Fitting a formal linear regression model

## Extract data for "direct" and "indirect" $f_4$-ratio

Filter data on the parameter setting of `rate_eur2afr == 0.2`, and
extract statistics for "direct" and "indirect" $f_4$-ratio statistics,
saving them in data frames `df_direct` and `df_indirect`:

```
df_direct <- filter( ... add requried code here ...)
df_indirect <- filter( ... add required code here ...)
```

## Run linear regression on "direct" vs "indirect" $f_4$-ratio

Run a linear model using the `lm()` function (help can be found under
`?lm`) to find a p-value of the statistical significance between trajectories
of "direct" and "indirect" $f_4$-ratio estimates:

```{r}
# computing a linear regresion on df_direct and df_indirect data
lm_direct <- lm(proportion ~ time, data = df_direct)
lm_indirect <- lm(proportion ~ time, data = df_indirect)
```

Run `summary()` on both objects with result of `lm()` to get a p-value.
Which statistic gives a statistically significant decline of Neanderthal
ancestry over time?

```{r}
# run summary on the lm fit of the direct f4-ratio
```

```{r}
# run summary on the lm fit of the indirect f4-ratio
```

# Conclusions

Here is where you could conclude what you found out in the course of the analysis.
Take home messages, pointers to follow up analyses in other later scripts and/or
Quarto notebooks, etc.